---
title: "Genomics QC Ancestry Report"
author: "GDC Data Team"
editor: visual
date: today
format:
  pdf:
    self-contained: true
editor: visual
execute:
  echo: false
  warning: false
  error: false
  message: false
  toc: true
project:
  output-dir: /home/gdc/shared/GDC_pipeline/results/
---

The following is an automated report describing the quality control (QC) processing of your dataset. detailed description of the QC steps taken.

```{r loadPackages}
#| message: false
#| echo: false
library(dplyr)
library(ggplot2)
library(gt)
library(forcats)
```

# Ancestry predictions

Ancestries are predicted after the initial round of QC steps. The principal components upon which the population is stratifed, are based off of the [genetic relatedness matrix (GRM)](https://zjuwhw.github.io/2021/08/20/GRM.html). The subjects are then projected onto the PC space and assigned ethnicities using k-means clustering based on either a standard genomic dataset (such as though genomes) are by the reported ancestries of the subject in the study. Figure @fig-ancestryPredictions the projection of each subject onto the first few PCs. They are colored by their reported ancestries.

The densities of ancestries based on the reference set are represent by the 10% contour lines of the 2d density plots with the observed populations presented as points over top of the densities shown in @fig-ancestryPredictions.


```{r importdata, echo=FALSE}
path_to_data="/home/gdc/shared/GDC_pipeline/results/Needed_files_for_report/SMILES_GSA/EAS/"
file_name="study.SMILES_GDA.unrelated.comm.popu"
#Data 
ancestry_prediction <- read.delim(paste0(path_to_data, file_name), header=FALSE)

```

<!-- ## Ancestry distribution -->

```{r datacleaning ancestryprediction}
# colnames(ancestry_prediction) = c("FID", "IID", "Ancestry", "Primary", "Statistic", "A1", "A2", "A3", "A4", "A5", "AFR", "AMR",
#                                   "EAS", "EUR", "SAS")
```


```{r Ancestry Distribution}
ggplot(data=ancestry_prediction, aes(x=Primary, y=Statistic, fill = Ancestry))+
  geom_area()
```

This is going to be a linked in png file
```{r ancestryPredictions}
#| label: fig-ancestryPredictions
#| fig-cap: "Ancestry predictions plotted over first decile contour for each ancestry."

data.frame(Ancestry = rep(c("Ancestry 1", "Ancestry 2"), each = 50),
           PC1 = c(rnorm(50, 0, 1), rnorm(50, 9, 1)),
           PC2 = c(rnorm(50, 0, 1), rnorm(50, 9, 1))) %>%
  ggplot(aes(x = PC1, y = PC2, color = Ancestry)) +
  geom_density_2d(breaks = c(0.05)) + 
  geom_point() +
  ggtitle("Ancestry Predictions")+
  theme_classic()
```
