---
title: "Lab: Visualizing Basic QC Outputs"
author: "Christian Coffman"
date: "2026-07-21"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

- **Full documentation**: [gdcgenomicsqc.readthedocs.io](https://gdcgenomicsqc.readthedocs.io/en/latest/)
- **QC diagnostics**: first check. Sample/variant missingness, AF distributions, HWE p-values reveal genotyping errors, poor DNA, batch effects before downstream analysis
- **Data are synthetic**: resampled from 1000 Genomes haplotypes. Expect textbook distributions:
  - Tight peaks near 0% missingness
  - U-shaped AF spectrum (many rare + many common variants)
  - Uniform HWE p-values (no population structure in synthetic sampling)

## Input

- **QC thresholds come from two places**:
  - Config `convertNfilt:` controls VCF-to-PGEN conversion filters:
    - `info_r2_min: 0.9`: minimum imputation R2, null to skip
    - `filter_pass: true`: keep only PASS variants
    - `qual_min: null`: no QUAL filter
  - Hard thresholds in `workflow/scripts/filterStandard.sh` and `workflow/rules/Standard_QC.smk`:
    - MAF 0.01, HWE 1e-6, missingness 0.02 (sample + variant)

```yaml
# sandboxToy.yaml lines 2-19
INPUT: "/scratch.global/GDC/toyData/allSubjects_hg38.bed"             # Single file = whole-genome mode (no {CHR})
OUT_DIR: "/scratch.global/GDC/r25outputs/toy"                         # Top-level output directory
chromosomes: [1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22]

convertNfilt:                                                          # Conversion-stage filters (VCF -> PGEN)
    info_r2_min: 0.9    # Minimum imputation R2; null to skip; VCF only
    filter_pass: true   # Keep only FILTER == PASS
    qual_min: null      # Minimum QUAL score; null = no filter
```

- See [QC Pipeline Tutorial](https://gdcgenomicsqc.readthedocs.io/en/latest/tutorial_qc_pipeline.html) for full reference

## Output

### Sample Missingness

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
# CHANGE THESE to match your config YAML
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy"
subset <- "full"

# Whole-genome mode: single-file outputs (no {CHR} suffix)
smiss <- read_tsv(file.path(OUT_DIR, subset, "initial.smiss"))
glimpse(smiss)

ggplot(smiss, aes(x = F_MISS * 100)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = c(2, 10), color = "red", linetype = "dashed") +
  labs(x = "Missing Genotypes (%)", y = "Number of Samples",
       title = "Sample Missingness Distribution") +
  theme_minimal()
```

### Variant Missingness

```{r, message=FALSE, warning=FALSE, error=FALSE}
vmiss <- read_tsv(file.path(OUT_DIR, subset, "initial.vmiss"))
glimpse(vmiss)

ggplot(vmiss, aes(x = F_MISS * 100)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7) +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed") +
  labs(x = "Missing Calls (%)", y = "Number of Variants",
       title = "Variant Missingness Distribution") +
  theme_minimal()
```

### Allele Frequency

```{r, message=FALSE, warning=FALSE, error=FALSE}
freq <- read_tsv(file.path(OUT_DIR, subset, "MAF_check.afreq"))
glimpse(freq)

freq %>%
  ggplot(aes(x = ALT_FREQS)) +
  geom_histogram(bins = 40, fill = "coral", alpha = 0.7) +
  labs(x = "Alternate Allele Frequency", y = "Count",
       title = "Allele Frequency Spectrum") +
  theme_minimal()
```

### Hardy-Weinberg Equilibrium

```{r, message=FALSE, warning=FALSE, error=FALSE}
hwe <- read_tsv(file.path(OUT_DIR, subset, "standardFilter.hardy"))
glimpse(hwe)

hwe %>%
  ggplot(aes(x = P)) +
  geom_histogram(bins = 50, fill = "purple", alpha = 0.7) +
  scale_x_log10() +
  labs(x = "P-value (log scale)", y = "Number of Variants",
       title = "HWE P-Value Distribution") +
  theme_minimal()
```

### MAF vs HWE

```{r, message=FALSE, warning=FALSE, error=FALSE}
freq %>%
  inner_join(hwe, by = c("#CHROM" = "#CHROM", "ID" = "ID")) %>%
  ggplot(aes(x = ALT_FREQS, y = -log10(P))) +
  geom_point(alpha = 0.3, size = 0.5) +
  labs(x = "Alternate Allele Frequency",
       y = expression(-log[10](HWE ~ P)),
       title = "MAF vs. HWE") +
  theme_minimal()
```

### Apply Thresholds

```{r, message=FALSE, warning=FALSE, error=FALSE}
thresholds <- list(geno = 0.02, mind = 0.02, maf = 0.01, hwe = 1e-6)

cat("Variants passing --geno 0.02:", vmiss %>% filter(F_MISS <= thresholds$geno) %>% nrow(), "\n")
cat("Variants passing --hwe 1e-6:", hwe %>% filter(P >= thresholds$hwe) %>% nrow(), "\n")
cat("Variants passing --maf 0.01:", freq %>% filter(ALT_FREQS >= thresholds$maf & ALT_FREQS <= (1 - thresholds$maf)) %>% nrow(), "\n")
cat("Samples passing --mind 0.02:", smiss %>% filter(F_MISS <= thresholds$mind) %>% nrow(), "\n")
```

### Per-Ancestry QC Diagnostics

- After ancestry classification, pipeline re-applies QC filters per ancestry group (EUR, AFR, AMR). Reveals ancestry-specific patterns masked in pooled analysis. Especially HWE: population structure in full sample produces spurious deviations that disappear when each group is tested separately
- A variant passing HWE in one population may deviate in another due to: population-specific genotyping artifacts, different AF affecting test power, or genuine selection pressure differences. Per-ancestry diagnostics distinguish these

```{r, message=FALSE, warning=FALSE, error=FALSE}
ancestries <- c("EUR", "AFR", "AMR", "EAS", "SAS")
anc_qc <- map_dfr(set_names(ancestries), function(a) {
  f <- file.path(OUT_DIR, a, "initial.smiss")
  if (file.exists(f)) {
    read_tsv(f, show_col_types = FALSE) |>
      summarise(
        N = n(),
        mean_miss = mean(F_MISS) * 100,
        pct_above_2pct = mean(F_MISS > 0.02) * 100
      ) |>
      mutate(ancestry = a, .before = 1)
  } else {
    NULL
  }
}, .id = NULL)

if (nrow(anc_qc) > 0) {
  print(anc_qc)

  ggplot(anc_qc, aes(x = ancestry, y = mean_miss, fill = ancestry)) +
    geom_col(alpha = 0.7) +
    labs(x = "Ancestry", y = "Mean Sample Missingness (%)",
         title = "Sample Missingness by Ancestry") +
    theme_minimal()
} else {
  cat("Per-ancestry QC files not found.\n")
  cat("Re-run with ancestry-specific QC enabled to see this section.\n")
}
```

Similarly, compare the number of variants passing HWE per ancestry:

```{r, message=FALSE, warning=FALSE, error=FALSE}
anc_hwe <- map_dfr(set_names(ancestries), function(a) {
  f <- file.path(OUT_DIR, a, "standardFilter.hardy")
  if (file.exists(f)) {
    read_tsv(f, show_col_types = FALSE) |>
      summarise(
        N = n(),
        pct_pass_hwe = mean(P >= 1e-6) * 100
      ) |>
      mutate(ancestry = a, .before = 1)
  } else {
    NULL
  }
}, .id = NULL)

if (nrow(anc_hwe) > 0) {
  print(anc_hwe)

  ggplot(anc_hwe, aes(x = ancestry, y = pct_pass_hwe, fill = ancestry)) +
    geom_col(alpha = 0.7) +
    labs(x = "Ancestry", y = "Variants Passing HWE (%)",
         title = "HWE Filtering Impact by Ancestry") +
    theme_minimal()
} else {
  cat("Per-ancestry HWE files not found.\n")
}
```

### Heterozygosity

```{r, message=FALSE, warning=FALSE, error=FALSE}
het <- read_tsv(file.path(OUT_DIR, subset, "heterozygosity.het"))
glimpse(het)

het %>%
  ggplot(aes(x = F)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  geom_vline(xintercept = mean(het$F) + c(-3, 3) * sd(het$F),
             color = "red", linetype = "dashed") +
  labs(x = "Inbreeding Coefficient (F)", y = "Number of Samples",
       title = "Heterozygosity (F Statistic) Distribution with ±3SD Thresholds") +
  theme_minimal()
```

- **±3SD lines**: default het outlier threshold based on the inbreeding coefficient `F`. Samples outside are flagged for removal. Tight cluster expected; wide spread or multiple modes suggests contamination or batch effects.

- **What to look for**: if one ancestry group shows higher missingness or HWE deviation, possible causes:
  - **Population-specific genotyping artifacts**: assay performs differently across ancestral backgrounds (probe bias, differential amplification)
  - **Admixture affecting HWE**: real deviations from HWE (not genotyping errors) but still trigger the filter
  - **Small sample size**: fewer samples = wider CIs, less power, more extreme p-values by chance

## Discussion

- **Why does the allele-frequency histogram show a U-shape?** Excess of both rare (new mutations, low-frequency segregating sites) and common (ancestral, high-frequency derived) variants. Expected allele frequency spectrum under neutral evolution with population growth.
- **If a dataset had a spike of HWE p-values near 0**, two possible explanations: (1) genotyping errors or batch effects causing systematic miscalls, or (2) population structure (multiple ancestries in sample, variants differ in frequency between groups). Per-ancestry diagnostics distinguish these.
- **Sample vs. variant missingness > 5%: which is more concerning?** High sample missingness = poor DNA/processing failure for that sample. Remove a few bad samples. High variant missingness across many samples = systematic assay problem (bad probe, poor genomic region). Affects every sample for that variant, harder to fix. Both > 5%? Check overlap: are poor variants concentrated in poor samples, or independent issues?

See the [QC Pipeline Tutorial](https://gdcgenomicsqc.readthedocs.io/en/latest/tutorial_qc_pipeline.html) for troubleshooting and deep dives into file formats and pipeline internals.
