---
title: "Lab: Visualizing Basic QC Outputs"
author: "Christian Coffman"
date: "2026-06-12"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

Before any downstream analysis, you need to know your data is clean. Per-sample and per-variant missingness, allele frequency distributions, and Hardy-Weinberg equilibrium p-values reveal genotyping errors, poor-quality DNA samples, and batch effects.

The data are **synthetic** — resampled from 1000 Genomes haplotypes — so the distributions should look textbook-perfect: tight peaks near 0% missingness, the classic U-shaped AF spectrum, and uniform HWE p-values.

## Input

The pipeline's QC thresholds are set partly in the config and partly hardcoded in the filtering scripts. The config controls the conversion-stage filters:

```yaml
# sandboxToy.yaml lines 2-19 (relevant keys)
INPUT: "/scratch.global/GDC/toyData/allSubjects_hg38_chr{CHR}.vcf.gz"  # Genotype input: VCF, PLINK1, or PLINK2 format
OUT_DIR: "/scratch.global/GDC/r25outputs/toy_chr"                      # Top-level output directory
chromosomes: [13, 14, 15, 16, 17, 18, 19, 20, 21, 22]                 # Chromosomes to process

convertNfilt:                                                          # Conversion-stage filters (VCF → PGEN)
    info_r2_min: 0.9    # Minimum imputation R² (INFO field); set null to skip; only applies to VCF
    filter_pass: true   # Keep only variants with FILTER == PASS; false = keep all
    qual_min: null      # Minimum QUAL score; null = no QUAL filter; numeric value e.g. 30
```

The hard MAF (0.01), HWE (1e-6), and missingness (0.02) thresholds are in `workflow/scripts/filterStandard.sh` and `workflow/rules/Standard_QC.smk`. The full pipeline reference is in the [QC Pipeline Tutorial](../tutorial_qc_pipeline.rst).

## Output

### Sample Missingness

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy"
subset <- "full"

smiss <- read_tsv(file.path(OUT_DIR, subset, "initial.smiss"))
glimpse(smiss)

ggplot(smiss, aes(x = F_MISS * 100)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = c(2, 10), color = "red", linetype = "dashed") +
  labs(x = "Missing Genotypes (%)", y = "Number of Samples",
       title = "Sample Missingness Distribution") +
  theme_minimal()
```

### Variant Missingness

```{r, message=FALSE, warning=FALSE, error=FALSE}
vmiss <- read_tsv(file.path(OUT_DIR, subset, "initial.vmiss"))
glimpse(vmiss)

ggplot(vmiss, aes(x = F_MISS * 100)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7) +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed") +
  labs(x = "Missing Calls (%)", y = "Number of Variants",
       title = "Variant Missingness Distribution") +
  theme_minimal()
```

### Allele Frequency

```{r, message=FALSE, warning=FALSE, error=FALSE}
freq <- read_tsv(file.path(OUT_DIR, subset, "MAF_check.afreq"))
glimpse(freq)

freq %>%
  ggplot(aes(x = ALT_FREQS)) +
  geom_histogram(bins = 40, fill = "coral", alpha = 0.7) +
  labs(x = "Alternate Allele Frequency", y = "Count",
       title = "Allele Frequency Spectrum") +
  theme_minimal()

freq %>%
  ggplot(aes(x = ALT_FREQS)) +
  geom_histogram(bins = 40, fill = "coral", alpha = 0.7) +
  facet_wrap(~`#CHROM`) +
  labs(x = "Alternate Allele Frequency", y = "Count",
       title = "Allele Frequency by Chromosome") +
  theme_minimal()
```

### Hardy-Weinberg Equilibrium

```{r, message=FALSE, warning=FALSE, error=FALSE}
hwe <- read_tsv(file.path(OUT_DIR, subset, "standardFilter.hardy"))
glimpse(hwe)

hwe %>%
  ggplot(aes(x = P)) +
  geom_histogram(bins = 50, fill = "purple", alpha = 0.7) +
  scale_x_log10() +
  labs(x = "P-value (log scale)", y = "Number of Variants",
       title = "HWE P-Value Distribution") +
  theme_minimal()
```

### MAF vs HWE

```{r, message=FALSE, warning=FALSE, error=FALSE}
freq %>%
  inner_join(hwe, by = c("#CHROM" = "#CHROM", "ID" = "ID")) %>%
  ggplot(aes(x = ALT_FREQS, y = -log10(P))) +
  geom_point(alpha = 0.3, size = 0.5) +
  labs(x = "Alternate Allele Frequency",
       y = expression(-log[10](HWE ~ P)),
       title = "MAF vs. HWE") +
  theme_minimal()
```

### Apply Thresholds

```{r, message=FALSE, warning=FALSE, error=FALSE}
thresholds <- list(geno = 0.02, mind = 0.02, maf = 0.01, hwe = 1e-6)

cat("Variants passing --geno 0.02:", vmiss %>% filter(F_MISS <= thresholds$geno) %>% nrow(), "\n")
cat("Variants passing --hwe 1e-6:", hwe %>% filter(P >= thresholds$hwe) %>% nrow(), "\n")
cat("Variants passing --maf 0.01:", freq %>% filter(ALT_FREQS >= thresholds$maf & ALT_FREQS <= (1 - thresholds$maf)) %>% nrow(), "\n")
cat("Samples passing --mind 0.02:", smiss %>% filter(F_MISS <= thresholds$mind) %>% nrow(), "\n")
```

## Discussion

1. **Why does the allele-frequency histogram show a U-shape?** What does that tell you about the genetic diversity of the 1KG reference panel?
2. **If a dataset had a spike of HWE p-values near 0**, what are two possible explanations?
3. **Sample vs. variant missingness**: which is more concerning when both exceed 5%?

See the [QC Pipeline Tutorial](../tutorial_qc_pipeline.rst) for troubleshooting and deep dives into file formats and pipeline internals.
