---
title: "Lab: Visualizing Global Ancestry Outputs"
author: "Christian Coffman"
date: "2026-07-22"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

- **A random forest trained on 1000 Genomes PCA coordinates** classifies each sample into one of five 1KG superpopulations (AFR, AMR, EAS, EUR, SAS). With synthetic data derived from the same reference, assignments should be near 100% accurate because the populations are cleanly separated in PCA space
- **The pipeline supports multiple latent spaces**: PCA (linear), UMAP (non-linear), VAE (variational autoencoder, probabilistic, handles missing data). Toggle with single config key (`ancestry.model`). Output file format unchanged across all three, so R code does not change

## Input

```yaml
# sandboxToy.yaml lines 82-85
ancestry:
    threshold: 0.8        # Minimum posterior probability for hard classification; samples below this -> "Other"
    model: "pca"           # Latent space for RF: "pca" (linear), "umap" (non-linear), "vae" (deep learning)
    ancestry_file: null    # Optional path to pre-computed ancestry labels; null = use pipeline classification
```

- **`model`**: selects latent space for RF. Options: `"pca"` (linear, fast), `"umap"` (non-linear, fine-scale structure), `"vae"` (deep learning, requires pre-computed VAE coordinates)
- **`threshold`**: minimum posterior probability for hard call. Below this, samples labeled "Other". Lower threshold = fewer "Other" labels, more misclassifications near ancestry boundaries
- **Switching models**: one-line config change (`model: "umap"`), re-run. Pipeline computes UMAP from PCA coordinates and trains new RF. Output file format unchanged, so R code does not change

## Output

### PCA Projection

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
# CHANGE THIS to match your OUT_DIR from the config YAML
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy_chr"
global_anc_dir <- file.path(OUT_DIR, "01-globalAncestry")

sample_pca <- read_tsv(file.path(global_anc_dir, "sampleRefPCscores.sscore")) |>
  select(IID = `#IID`, PC1 = PC1_AVG, PC2 = PC2_AVG) |>
  mutate(sim_ancestry = str_sub(IID, 1, 3))

ref_pca <- read_tsv(file.path(global_anc_dir, "refRefPCscores.sscore")) |>
  select(IID = `#IID`, PC1 = PC1_AVG, PC2 = PC2_AVG)

labels <- read_table("/scratch.global/GDC/referenceData/1000G_highcoverage/population.txt") |>
  select(IID = SampleID, POP = Superpopulation)
ref_pca <- ref_pca |> inner_join(labels, by = "IID")

ggplot() +
  geom_density_2d(data = ref_pca, aes(x = PC1, y = PC2, color = POP), alpha = 0.4) +
  geom_point(data = sample_pca, aes(x = PC1, y = PC2, color = sim_ancestry),
             size = 0.8, alpha = 0.3) +
  labs(x = "PC1", y = "PC2", color = "Ancestry",
       title = "Study Samples in 1KG Reference PCA Space") +
  theme_minimal()
```

### Classification Probabilities

```{r, message=FALSE, warning=FALSE, error=FALSE}
class_prob <- read_tsv(file.path(global_anc_dir, "classificationProbabilities.tsv"))
class <- read_tsv(file.path(global_anc_dir, "ancestry_classifications.tsv"))

anc_data <- class_prob |>
  left_join(class, by = "IID") |>
  mutate(ancestry = str_sub(IID, 1, 3))

anc_data |>
  pivot_longer(
    cols = ends_with(c("_AFR", "_AMR", "_EAS", "_EUR", "_SAS")),
    names_to = "supposed_ancestry",
    values_to = "classification_prob"
  ) |>
  slice_head(n = 50, by = ancestry) |>
  ggplot(aes(x = reorder(IID, classification_prob), y = classification_prob, fill = supposed_ancestry)) +
  geom_col(width = 1) +
  labs(x = "Sample", y = "Classification Probability", fill = "Ancestry",
       title = "Per-Sample Classification Probabilities") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~ancestry, scales = "free_x")
```

### Confusion Matrix & Agreement

```{r, message=FALSE, warning=FALSE, error=FALSE}
table(anc_data$ancestry, anc_data$pca_predicted)

agreement <- anc_data |>
  mutate(pca_truth_agree = pca_predicted == ancestry)
cat("PCA vs truth agreement:", mean(agreement$pca_truth_agree, na.rm = TRUE) * 100, "%\n")
```

### Confidence by Ancestry

```{r, message=FALSE, warning=FALSE, error=FALSE}
anc_data |>
  select(IID, ancestry, ends_with("_confidence")) |>
  pivot_longer(cols = -c(IID, ancestry), names_to = "method", values_to = "confidence") |>
  mutate(method = str_remove(method, "_confidence")) |>
  ggplot(aes(x = ancestry, y = confidence, fill = ancestry)) +
  geom_violin(alpha = 0.7) +
  labs(x = "Ancestry", y = "Confidence",
       title = "Classification Confidence by Ancestry") +
  theme_minimal()
```

## Going Further: PCA vs UMAP

- **All model probabilities in one file**: `classificationProbabilities.tsv` has both `pca_AFR` and `umap_AFR` columns (and others). Even with `model: "pca"`, you can compare both methods side-by-side
- **No re-run needed**: the data to compare models is already there. Code below extracts EUR probabilities from PCA vs UMAP for the same samples

**Compare PCA and UMAP classification probabilities:**

```{r, message=FALSE, warning=FALSE, error=FALSE}
# class_prob already has both pca_ and umap_ columns if UMAP was provided as input
# Compare EUR probabilities from both models
class_prob |>
  select(IID, pca_EUR, umap_EUR) |>
  pivot_longer(-IID, names_to = "model", values_to = "eur_prob") |>
  ggplot(aes(x = eur_prob, fill = model)) +
  geom_density(alpha = 0.5) +
  labs(x = "EUR Probability", y = "Density", fill = "Model",
       title = "PCA vs UMAP: EUR Probability Distribution") +
  theme_minimal()
```

- **Model comparison:**
  - **PCA** (linear): separates populations along major genetic variation axes. Closely related populations (EUR vs SAS) may have overlapping probabilities because their separation is in higher PCs. Efficient, well-understood, default for well-separated continental populations
  - **UMAP** (non-linear): captures fine-scale structure PCA compresses into higher PCs by preserving local neighborhoods. Higher confidence near population boundaries, but can overfit to noise in small datasets
  - **VAE** (deep learning): probabilistic latent space with uncertainty estimates. Handles missing data naturally, but requires much more compute and tuning than PCA/UMAP
- **When each model wins:**
  - Well-separated continental populations: PCA sufficient. Clean linear separation
  - Fine-scale structure (Finnish vs British EUR, Gujarati vs Punjabi SAS): UMAP finds patterns PCA misses by preserving local neighborhoods
  - Substantial missing data (ancient DNA, low-coverage sequencing): VAE more robust, models observation-level uncertainty instead of requiring complete cases

## Going Further: Local Ancestry Inference

- **Global ancestry = one label per sample**. RF gives single posterior probability vector per individual (e.g., 80% AFR, 20% EUR). But an admixed genome is a mosaic: entire chromosomes may be from one population or the other
- **Local ancestry = ancestry per position**. ShapeIt4 phases haplotypes, RFMix deconvolves which population each segment came from. Output: per-variant, per-haplotype ancestry calls along every chromosome, revealing recombination breakpoints
- **When local ancestry matters:**
  - **AFR/EUR admixed sample**: one haplotype entirely EUR on chr1, entirely AFR on chr2. Different genes sit in different LD and MAF contexts depending on ancestral background
  - **Heritability and PRS assuming uniform ancestry** miss regional effects. Variant in AFR-derived block has different effect size + LD tagging than same variant in EUR-derived block. PRS trained on EUR summary stats performs poorly on AFR-derived segments
  - **Fine-mapping in admixed populations** needs local ancestry to distinguish truly causal variants from those correlated with a nearby ancestry switch (ancestry confounding, which produces false association signals)

### Discussion: Local ancestry

- **Many short ancestry tracts: real admixture or phasing artifact?** Tract length distribution is key: short even tracts across all chromosomes = recent admixture (many recombination events since admixture). Long tracts on specific chromosomes = phasing artifact (phaser incorrectly switches haplotypes). Validate by checking tract-length distribution against admixture model (exponential decay with generations) and comparing results across different phasing parameters.
- **Most samples assigned single local ancestry: RFMix failure or genuinely unadmixed?** Check reference panel. EUR-only panel on EUR samples: single ancestry expected, RFMix correct. Multi-population panel but everyone assigned one ancestry: reference populations too similar, or samples unadmixed but from population not well-represented in reference. Cross-check with global ancestry classification.
- **Do RFMix genome-wide proportions agree with RF posterior probabilities?** Broadly yes for well-separated populations. RFMix EUR proportion should roughly match RF EUR posterior. Detail differs: RFMix per-position estimates are noisy at ancestry transitions, RF posterior is a sample-level summary influenced by most informative PCs. Systematic disagreement (RFMix 50% EUR vs RF 80% EUR) suggests bias in one method.
- **How does local ancestry improve PRS in admixed individuals?** Variant in AFR-derived block: LD matches AFR reference, not EUR. Use AFR LD for clumping/LDpred2 for that variant, EUR LD for EUR-derived blocks. Ancestry-aware PRS matches LD reference to local ancestry context per variant, substantially improving prediction accuracy.
- **Phasing accuracy in recently admixed populations?** ShapeIt4 uses reference panel. Panel mostly EUR + sample AFR/EUR admixed: EUR haplotypes phase more accurately (many EUR haplotypes in reference with similar LD). AFR haplotypes mis-phased more often, especially at ancestry transition points where LD structure switches abruptly. Phasing errors propagate to RFMix: a phasing switch error creates a spurious ancestry switch in RFMix output.

See the [Multi-Ancestry PRS Tutorial](../tutorial_prs_multi.rst) for local ancestry parameter tuning and troubleshooting.

## Discussion: Global Classification

- **Why near-perfect classification with synthetic data?** Resampled from 1000 Genomes haplotypes, same population structure in study and reference. PCA projects synthetic samples directly onto source population clusters. In real data, off-diagonal confusion matrix entries come from: genuine admixture (mixed ancestry falls between clusters), population structure within a superpopulation (Finnish vs British both EUR but may classify differently), and genotyping artifacts shifting PCA coordinates.
- **How does `threshold` affect the confusion matrix?** Sets minimum posterior probability for hard classification. Below threshold (e.g., 0.8), samples labeled "Other". Lower threshold = fewer "Other" labels, more misclassifications at ancestry boundaries. Higher threshold = more "Other" labels, higher precision for classified samples. Use case determines optimal: clean ancestry strata for heritability -> high threshold. Classify every sample for PRS -> lower threshold, accept boundary misclassifications.
- **UMAP 99.5% vs PCA 98% agreement: meaningful?** 1.5pp difference could be real or noise. Test: (1) bootstrap confidence intervals for both estimates, overlapping CIs = not significant. (2) Which samples are misclassified? UMAP correctly classifying boundary samples PCA misses = qualitatively meaningful. (3) Is improvement concentrated in specific populations? (UMAP may improve AMR classification, which is admixed between EUR/AFR/SAS in PCA space). A small improvement may be practically irrelevant if downstream conclusions (heritability estimates) do not change.
