---
title: "Lab: Visualizing Global Ancestry Outputs"
author: "Christian Coffman"
date: "2026-07-22"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

A random forest trained on 1000 Genomes PCA coordinates classifies each sample into a 1KG superpopulation (AFR, AMR, EAS, EUR, SAS). With synthetic data derived from the same reference, assignments should be near 100% accurate — the populations are cleanly separated in PCA space.

The pipeline supports multiple latent spaces: PCA (linear), UMAP (non-linear), and VAE (variational autoencoder). You toggle which model is used with a single config value — no code changes needed.

## Input

```yaml
# sandboxToy.yaml lines 82-85
ancestry:
    threshold: 0.8        # Minimum posterior probability for a hard classification; samples below this → "Other"
    model: "pca"           # Latent space for RF: "pca" (linear, fast), "umap" (non-linear, fine-scale), "vae" (deep learning)
    ancestry_file: null    # Optional path to pre-computed ancestry labels; null = use pipeline classification
```

`model` selects the latent space: `"pca"` (linear, fast, well-understood), `"umap"` (non-linear, captures fine-scale structure), or `"vae"` (deep learning, requires VAE coordinates). `threshold` is the minimum posterior probability for a hard call — samples below it are labeled "Other".

**To switch to UMAP**, change `model: "umap"` and re-run. The pipeline automatically computes the UMAP embedding from the PCA coordinates and trains a new random forest. The output files have the same format — your R code does not change.

## Output

### PCA Projection

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
# CHANGE THIS to match your OUT_DIR from the config YAML
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy_chr"
global_anc_dir <- file.path(OUT_DIR, "01-globalAncestry")

sample_pca <- read_tsv(file.path(global_anc_dir, "sampleRefPCscores.sscore")) |>
  select(IID = `#IID`, PC1 = PC1_AVG, PC2 = PC2_AVG) |>
  mutate(sim_ancestry = str_sub(IID, 1, 3))

ref_pca <- read_tsv(file.path(global_anc_dir, "refRefPCscores.sscore")) |>
  select(IID = `#IID`, PC1 = PC1_AVG, PC2 = PC2_AVG)

labels <- read_table("/scratch.global/GDC/referenceData/1000G_highcoverage/population.txt") |>
  select(IID = SampleID, POP = Superpopulation)
ref_pca <- ref_pca |> inner_join(labels, by = "IID")

ggplot() +
  geom_density_2d(data = ref_pca, aes(x = PC1, y = PC2, color = POP), alpha = 0.4) +
  geom_point(data = sample_pca, aes(x = PC1, y = PC2, color = sim_ancestry),
             size = 0.8, alpha = 0.3) +
  labs(x = "PC1", y = "PC2", color = "Ancestry",
       title = "Study Samples in 1KG Reference PCA Space") +
  theme_minimal()
```

### Classification Probabilities

```{r, message=FALSE, warning=FALSE, error=FALSE}
class_prob <- read_tsv(file.path(global_anc_dir, "classificationProbabilities.tsv"))
class <- read_tsv(file.path(global_anc_dir, "ancestry_classifications.tsv"))

anc_data <- class_prob |>
  left_join(class, by = "IID") |>
  mutate(ancestry = str_sub(IID, 1, 3))

anc_data |>
  pivot_longer(
    cols = ends_with(c("_AFR", "_AMR", "_EAS", "_EUR", "_SAS")),
    names_to = "supposed_ancestry",
    values_to = "classification_prob"
  ) |>
  slice_head(n = 50, by = ancestry) |>
  ggplot(aes(x = reorder(IID, classification_prob), y = classification_prob, fill = supposed_ancestry)) +
  geom_col(width = 1) +
  labs(x = "Sample", y = "Classification Probability", fill = "Ancestry",
       title = "Per-Sample Classification Probabilities") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~ancestry, scales = "free_x")
```

### Confusion Matrix & Agreement

```{r, message=FALSE, warning=FALSE, error=FALSE}
table(anc_data$ancestry, anc_data$pca_predicted)

agreement <- anc_data |>
  mutate(pca_truth_agree = pca_predicted == ancestry)
cat("PCA vs truth agreement:", mean(agreement$pca_truth_agree, na.rm = TRUE) * 100, "%\n")
```

### Confidence by Ancestry

```{r, message=FALSE, warning=FALSE, error=FALSE}
anc_data |>
  select(IID, ancestry, ends_with("_confidence")) |>
  pivot_longer(cols = -c(IID, ancestry), names_to = "method", values_to = "confidence") |>
  mutate(method = str_remove(method, "_confidence")) |>
  ggplot(aes(x = ancestry, y = confidence, fill = ancestry)) +
  geom_violin(alpha = 0.7) +
  labs(x = "Ancestry", y = "Confidence",
       title = "Classification Confidence by Ancestry") +
  theme_minimal()
```

## Going Further: PCA vs UMAP

The pipeline outputs probability columns for **every model** in a single `classificationProbabilities.tsv`. So even with `model: "pca"`, if UMAP coordinates were provided as input, both `pca_AFR` and `umap_AFR` columns exist. You already have the data to compare them.

**To compare PCA and UMAP classification probabilities for the same samples:**

```{r, message=FALSE, warning=FALSE, error=FALSE}
# class_prob already has both pca_ and umap_ columns if UMAP was provided as input
# Compare EUR probabilities from both models
class_prob |>
  select(IID, pca_EUR, umap_EUR) |>
  pivot_longer(-IID, names_to = "model", values_to = "eur_prob") |>
  ggplot(aes(x = eur_prob, fill = model)) +
  geom_density(alpha = 0.5) +
  labs(x = "EUR Probability", y = "Density", fill = "Model",
       title = "PCA vs UMAP: EUR Probability Distribution") +
  theme_minimal()
```

**What to look for:**
- **PCA** (linear) separates populations that differ along the major axes of genetic variation. Closely related populations (e.g., EUR and SAS) may show overlapping probability distributions.
- **UMAP** (non-linear) can capture fine-scale structure that PCA compresses into higher PCs. It may assign higher confidence to samples near population boundaries, but it can also overfit to noise in small datasets.
- **VAE** (deep learning) learns a probabilistic latent space. It handles missing data naturally but requires more compute and tuning.

**When would one outperform another?** For well-separated continental populations, PCA is usually sufficient — the linear separation is clean. For fine-scale structure (e.g., distinguishing Finnish from British within EUR), UMAP's non-linear embedding can find patterns PCA misses. For datasets with substantial missingness, VAE's probabilistic approach is more robust.

## Going Further: Local Ancestry Inference

The global ancestry classification above assigns a single ancestry per *sample*. Local ancestry goes a step further: it assigns an ancestry to each *position along each chromosome*, deconvolving the two haplotypes. The pipeline supports this with ShapeIt4 phasing + RFMix when `localAncestry.RFMIX: true` is set in the config.

**When does local ancestry matter?**
- A sample classified as AFR/EUR admixed may have long tracts of EUR origin and long tracts of AFR origin — different genes fall in different ancestry contexts.
- Heritability and PRS methods that assume uniform ancestry per sample will miss these regional effects. A variant in an AFR-derived block may have a different effect size than the same variant in an EUR-derived block.
- Local ancestry is essential for fine-mapping in admixed populations and for building ancestry-aware PRS.

### Discussion: Local ancestry

1. **Samples with many short ancestry tracts — is this real admixture or a phasing artifact?** How would you distinguish them? (Hint: tract length distribution; short tracts = many generations since admixture.)
2. **If most samples are assigned a single local ancestry**, does that mean RFMix failed or that the population is genuinely unadmixed? Check the reference panel composition.
3. **How do the global ancestry proportions from a local ancestry analysis (RFMix) compare to the RF posterior probabilities from the classification step?** Would you expect them to agree?
4. **How could local ancestry be used to improve PRS in admixed individuals?** If a variant sits in an AFR-derived block, its LD structure matches AFR reference panels, not EUR — how would you choose the LD reference for clumping or LDpred2?
5. **What happens to phasing accuracy in recently admixed populations?** ShapeIt4 uses a reference panel — if the panel is mostly EUR but your sample is AFR/EUR admixed, which haplotypes will be phased more accurately? How could phasing errors propagate into RFMix ancestry calls?

See the [Multi-Ancestry PRS Tutorial](../tutorial_prs_multi.rst) for local ancestry parameter tuning and troubleshooting.

## Discussion: Global Classification

1. **Why should you expect near-perfect classification with synthetic data?** What would cause off-diagonal entries in a real dataset?
2. **How does the `threshold` config value affect the confusion matrix?** What happens to samples whose max probability is below 0.8?
3. **If UMAP gives 99.5% agreement and PCA gives 98%, is that a meaningful improvement?** How would you test whether the difference is real or noise?
