---
title: "Lab: Visualizing SNP Heritability Outputs"
author: "Christian Coffman"
date: "2026-06-12"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

SNP heritability (h²) estimates the proportion of phenotypic variance explained by genetic variants. The pipeline uses MASH (or GCTA) to estimate h² within each ancestry subset using a GRM computed from LD-pruned, crossmapped variants. This lab compares estimates between **EUR** and **AFR** subsets.

The genotypes are **synthetic** — resampled from 1000 Genomes haplotypes and available on Dataverse. The phenotype is simulated with a **known heritability of 0.4**. This gives you a ground-truth target: both ancestry estimates should be near 0.4.

## Goals & Expectations

Frame your exploration around these questions:

1. **Heritability recovery** — Do the EUR and AFR h² estimates center around the true value of 0.4?
2. **Cross-ancestry comparison** — Are the estimates similar, or does one ancestry show bias?
3. **Statistical significance** — Are both estimates significantly different from zero (p < 0.05)?
4. **GRM structure** — Does the genetic relationship matrix look clean after KING filtering?

Given the **simulated phenotype** with h² = 0.4, both estimates should be near 0.4. If one is systematically biased, that suggests an ancestry-specific issue in the GRM construction or the estimation model.

## Inputs & Outputs

- **Inputs:** LD-pruned, crossmapped PGEN (`f1.b38.ldpruned.*`); phenotype file (simulated with h² = 0.4); covariate file (age, sex, PCs); GRM from GCTA.
- **Outputs:** `mash_output.csv` (h² estimates with SE and p-value for each ancestry).
- **Pipeline direction:** These estimates quantify how much of the trait variance is genetic. They inform power calculations for GWAS and PRS modeling.

## Upstream Pipeline Steps

Heritability estimation is sensitive to variant quality, LD, and sample structure. The pipeline runs these steps first:

1. **Initial QC** — Removes variants and samples with high missingness that would bias the GRM.
2. **Standard QC** (`filterStandard.sh` → `Standard_QC.smk`) — Applies MAF (>1%), HWE (p > 1e-6), and heterozygosity filters. **Low-MAF variants inflate heritability** because their effects are poorly estimated; HWE deviations suggest genotyping error rather than true genetic signal.
3. **LD pruning** (`concatLdPruned`) — The GRM computed by GCTA assumes **independent genetic markers**. Including correlated SNPs would double-count shared signal and inflate h² estimates.
4. **Crossmap** (`crossmapLdPrunedToB38`) — Ensures all variants map to the same genome build for consistent GRM computation.
5. **GRM computation** (`checkRelatednessExtractUnrelated` via `gcta64 --make-grm`) — Computes the genetic relationship matrix from the LD-pruned variants.
6. **Relatedness filtering** (`king` → `unrelated_ids.txt`) — Removes one individual from each pair with pi-hat > 0.0884. **Related samples inflate heritability** because they share both phenotype and genotype more than expected by chance.
7. **Phenotype simulation** (external / pre-pipeline) — A phenotype was simulated with **known h² = 0.4** and added to the synthetic genotype data. This provides a ground truth that real studies lack.

**Why independence matters here:** The GRM is a covariance matrix among individuals. If related individuals are included, the off-diagonal GRM elements are non-zero due to recent shared ancestry, not due to the genetic architecture of the trait. This biases h² estimates upward (confounds population structure with heritability).

## Configuration

Heritability estimation uses the `snpHerit` key in `config.yaml`. Here is a mock excerpt showing surrounding context:

```yaml
# config.yaml (mock excerpt — heritability-related keys only)
OUT_DIR: "/scratch/.../output"
REF: "/scratch/.../1000G_reference"

thresholds:
  maf: 0.01
  hwe: 1e-6

ancestry:
  model: pca

```

- The GRM and PCA paths are **auto-derived** from the ancestry wildcard — no `grm_prefix` or `pca_input` config needed. The pipeline concatenates per-chromosome LD-pruned files (`concatLdPruned`), crossmaps them to GRCh38, computes the GRM via `checkRelatednessExtractUnrelated`, then estimates h² per ancestry with MASH.
- The phenotype was **simulated externally** with known h² = 0.4. In a real study, it would come from clinical measurements.

The `mash_output.csv` files below contain the h² estimates for the **EUR** and **AFR** ancestry subsets.

## Before You Begin

- **Required files:** `mash_output.csv` in `{herit_dir}/EUR/03-snpHeritability/` and `{herit_dir}/AFR/03-snpHeritability/`; GRM binary files for the heatmap section
- **Disk space:** < 1 MB for CSV; GRM files can be 100 MB–2 GB
- **Expected render time:** < 5 seconds for CSV; 10–30 seconds for GRM heatmap (depends on n_samples)
- **Known issue:** If `mash_output.csv` is empty, the MASH solver may have failed to converge. Check the `run_snpHerit` log.
- **Reference:** The phenotype was simulated externally with h² = 0.4 to provide a known ground truth.

## Setup

The code chunk below adds the conda R library path so that pre-installed packages are found without a local `install.packages()`. The environment at `/scratch.global/R25_files/r25_r_env/` is shared across users and contains all packages used in these labs:

- **tidyverse** (ggplot2, dplyr, tidyr, readr, purrr, stringr, forcats, tibble) — data manipulation and plotting
- **data.table** — fast file reading via `fread()`

```{r setup}
conda_env <- '/scratch.global/R25_files/r25_r_env/lib/R/library'
.libPaths(unique(c(.Library.site, .Library)))
.libPaths(c(conda_env, .libPaths()))
library(tidyverse)

OUT_DIR <- "/path/to/your/pipeline/output"
herit_dir <- file.path(OUT_DIR, "03-snpHeritability")
```

## Load Heritability Estimates

**Files:** `{herit_dir}/{subset}/03-snpHeritability/mash_output.csv` for EUR and AFR.

**Format:** CSV with columns for ancestry, method, h² estimate, standard error (SE), and p-value.

**Pipeline step:** Generated by the `run_snpHerit` rule. It builds a GRM (`checkRelatednessExtractUnrelated` → `concatLdPruned` → `crossmapLdPrunedToB38`), then runs MASH to estimate variance components per ancestry.

```{r}
subsets <- c("EUR", "AFR")
herit <- map_dfr(subsets, function(s) {
  read_csv(file.path(herit_dir, s, "03-snpHeritability", "mash_output.csv"), show_col_types = FALSE)
})
glimpse(herit)
```

## Calculate 95% Confidence Intervals

Compute approximate CIs using the normal approximation (estimate ± 1.96 × SE).

```{r}
herit <- herit %>%
  mutate(
    lower_ci = h2 - 1.96 * SE,
    upper_ci = h2 + 1.96 * SE
  )
print(herit)
```

**Expectation:** The true h² = 0.4 should fall within the 95% CI for both ancestries.

## Point Estimate Comparison

Compare EUR and AFR h² estimates side by side. The dashed line marks the ground-truth value of 0.4.

```{r}
ggplot(herit, aes(x = ancestry, y = h2, color = ancestry)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  geom_hline(yintercept = 0.4, linetype = "dashed", color = "gray50", linewidth = 1) +
  labs(
    title = "SNP Heritability Estimates by Ancestry",
    subtitle = "Ground truth h² = 0.4 (dashed line)",
    x = "Ancestry",
    y = expression("Heritability (h"^2*")")
  ) +
  ylim(0, max(1, max(herit$upper_ci) * 1.2)) +
  theme_minimal()
```

## Statistical Significance

Are the h² estimates significantly different from zero (p < 0.05)?

```{r}
herit %>%
  filter(p_value < 0.05)
```

**Expectation:** Because h² = 0.4 is substantially above zero, both estimates should be highly significant. A non-significant result would indicate insufficient sample size or excessive standard error for that ancestry.

## GRM Heatmap

Visualize the genetic relationship matrix to confirm that KING filtering removed related samples. A well-filtered GRM should be near-diagonal.

```{r}
grm_id_path <- file.path(OUT_DIR, "relatedness", "f1.b38.ldpruned.unrelated.grm.id")
grm_bin_path <- file.path(OUT_DIR, "relatedness", "f1.b38.ldpruned.unrelated.grm.bin")
grm_ids <- read_tsv(grm_id_path, col_names = c("FID", "IID"))
n_ids <- nrow(grm_ids)
grm_bin <- readBin(grm_bin_path, what = "numeric", n = n_ids * (n_ids + 1) / 2)

grm_mat <- matrix(0, n_ids, n_ids)
idx <- 1
for (i in 1:n_ids) {
  for (j in 1:i) {
    grm_mat[i, j] <- grm_bin[idx]
    grm_mat[j, i] <- grm_bin[idx]
    idx <- idx + 1
  }
}

n_subsample <- min(200, n_ids)
set.seed(42)
subsample_idx <- sample(seq_len(n_ids), n_subsample)

grm_sub <- grm_mat[subsample_idx, subsample_idx]

grm_sub %>%
  as_tibble() %>%
  rowid_to_column("row") %>%
  pivot_longer(cols = -row, names_to = "col_str", values_to = "GRM") %>%
  mutate(col = as.integer(str_remove(col_str, "V"))) %>%
  ggplot(aes(x = col, y = row, fill = GRM)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Genetic Relationship Matrix (Subset)",
    subtitle = paste0(n_subsample, " samples — diagonal should dominate after KING filtering"),
    x = "Sample",
    y = "Sample"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank())
```

**Expectation:** The diagonal (self-relationship = ~1) should dominate. Off-diagonal values should cluster near 0. If you see substantial off-diagonal structure, KING filtering may not have been aggressive enough, or the GRM was computed before unrelated extraction.

## Check Your Understanding

1. **The ground-truth h² is 0.4, but the pipeline estimates 0.35.** Is this bias or just sampling variation? How would you decide?
2. **Look at your GRM heatmap. Are there any off-diagonal samples with GRM > 0.05?** What would you do with them if you found them?
3. **Why does LD pruning matter more for heritability estimation than for global ancestry classification?** (Hint: think about the GRM.)
4. **If EUR and AFR h² differ**, what factors (sample size, LD structure, MAF distribution) could drive the difference? Are these genuine biological differences or technical artifacts?

## Troubleshooting

| Problem | Likely Cause | Solution |
|---------|-------------|----------|
| `mash_output.csv` is empty | MASH failed to converge | Check `run_snpHerit` log; try increasing EM iterations or check for collinear covariates |
| GRM heatmap shows strong off-diagonal structure | KING filtering was skipped or pi-hat threshold too lenient | Verify `relatedness: king: true` in config; check the `unrelated_ids.txt` file size |
| h² estimate > 1 (or negative) | REML estimate on boundary; small sample size | With N < 100, REML is unreliable; interpret h² cautiously |
| h² estimate far from 0.4 | Possible GRM or phenotype issue | Check that the phenotype was simulated with h² = 0.4 and that LD pruning / crossmap completed successfully |

## Going Further

- **Power analysis**: Given your estimated h² and sample size per ancestry, calculate the minimum detectable effect size for a GWAS at 80% power (using `pwr` package in R).
- **Compare GCTA-REML to MASH**: If the config runs both methods, compute the correlation between their h² estimates. Which has smaller SEs?
- **Simulate a null phenotype**: Generate a phenotype with h² = 0, re-run through heritability estimation, and verify that the pipeline correctly estimates h² ≈ 0 with non-significant p-values.
