---
title: "Lab: Visualizing Single-Ancestry PRS Outputs"
author: "Christian Coffman"
date: "2026-06-12"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

Polygenic risk scores (PRS) aggregate the effects of many genetic variants into a single score that predicts phenotype. The pipeline runs PRSice-2 (clumping + thresholding) and LDpred2 (Bayesian point-normal prior). With a phenotype simulated at h² = 0.4, the PRS should capture a meaningful fraction of that signal.

## Input

```yaml
# sandboxToy.yaml lines 139-148
prsMethods:
  single_ct:          # Clumping + Thresholding (baseline); fast, no LD adjustment
    enabled: false
  single_prsice:      # PRSice-2: automated threshold scan, picks best R²; good default
    enabled: true
  single_prscs:       # PRS-CS: Bayesian regression with continuous shrinkage prior; requires external LD reference
    enabled: false
  single_ldpred2:     # LDpred2: Bayesian point-normal + auto model; accounts for LD, good with GWAS summary stats
    enabled: true
```

Only methods with `enabled: true` run. PRSice-2 tests multiple p-value thresholds and picks the best; LDpred2 scans a grid of prior parameters. For the full PRS pipeline reference, see the [PRS Tutorial](../tutorial_prs.rst).

## Output

### PRSice-2 Scores

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy_chr"
anc1 <- "AFR"; anc2 <- "EUR"
prs_out_dir <- file.path(OUT_DIR, "prs_inputs", paste0(anc1, "_", anc2))

prs_scores <- read_tsv(file.path(prs_out_dir, "single_prsice", "best_pRS.prs"),
                        col_names = c("FID", "IID", "PRS"))
glimpse(prs_scores)

ggplot(prs_scores, aes(x = PRS)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(x = "Polygenic Risk Score", y = "Number of Samples",
       title = "PRS Score Distribution (PRSice-2)") +
  theme_minimal()

prsice_summary <- read_csv(file.path(prs_out_dir, "single_prsice", "summary.csv"))
glimpse(prsice_summary)

prsice_summary |>
  ggplot(aes(x = Threshold, y = R.square)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  labs(x = "P-value Threshold", y = expression(R^2),
       title = "PRSice-2 Threshold Scan") +
  theme_minimal()
```

### LDpred2 Scores

```{r, message=FALSE, warning=FALSE, error=FALSE}
ldpred2_scores <- read_tsv(file.path(prs_out_dir, "single_ldpred2", "prs_scores.tsv"))
glimpse(ldpred2_scores)

ldpred2_scores |>
  select(IID, contains("auto")) |>
  pivot_longer(-IID, names_to = "model", values_to = "PRS") |>
  ggplot(aes(x = PRS, fill = model)) +
  geom_density(alpha = 0.5) +
  labs(x = "PRS", y = "Density", fill = "Model",
       title = "LDpred2 Auto Model Scores") +
  theme_minimal()
```

## Discussion

1. **Given h² = 0.4, what fraction of that heritability does the best PRS capture?** (Look at the R² in summary.csv.)
2. **Why would cross-ancestry transfer (training in AFR, testing in EUR) typically give lower R²?** What biological factors drive the drop?
3. **PRSice-2 vs. LDpred2 — which method gives higher R²?** Why might the Bayesian approach outperform C+T?

See the [PRS Tutorial](../tutorial_prs.rst) for cross-ancestry transfer exercises, multi-ancestry metaPRS comparison, and null simulations.
