---
title: "Lab: Visualizing and Working with Single-Ancestry PRS Outputs"
author: "GDCGenomicsQC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab: Visualizing and Working with Single-Ancestry PRS Outputs (R + Tidyverse)

This lab covers interacting with output files from the **Single-Ancestry PRS Pipeline** using R and tidyverse. You will work with PRS scores, performance metrics (R², AUC), and compare methods.

**Estimated time**: 1 hour

**Prerequisites**:
- Completed the PRS tutorial (outputs in `OUT_DIR/prs_inputs/{ANC1}_{ANC2}/`)
- R (≥4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

---

# Setup: Load Libraries and Set Paths

```{r load-libraries}
library(tidyverse)
OUT_DIR <- "/path/to/your/pipeline/output"
# PRS outputs are in OUT_DIR/prs_inputs/{ANC1}_{ANC2}/
anc1 <- "AFR"  # First ancestry
anc2 <- "EUR"  # Second ancestry
prs_out_dir <- file.path(OUT_DIR, "prs_inputs", paste0(anc1, "_", anc2))
```

---

# Section 1: Single-Ancestry PRS Output Files Reference

| File Path | Description |
|----------|------------|
| `single_prsice/` | PRSice2 output (best_pRS.prs, summary.csv) |
| `single_ldpred2/` | LDPred2 output (prs_scores.tsv) |
| `method_runs/{method}/` | Per-method output directories |
| `*.done` | Completion markers for each method |

---

# Section 2: Load and Inspect PRS Data

```{r load-data}
# Load PRS scores for PRSice2
prsice_score_path <- file.path(prs_out_dir, "single_prsice", "best_pRS.prs")
if (file.exists(prsice_score_path)) {
  prs_scores <- read_tsv(prsice_score_path, col_names = c("FID", "IID", "PRS"))
  glimpse(prs_scores)
}

# Load PRSice2 summary
prsice_summary_path <- file.path(prs_out_dir, "single_prsice", "summary.csv")
if (file.exists(prsice_summary_path)) {
  prsice_summary <- read_csv(prsice_summary_path)
  glimpse(prsice_summary)
}

# Load LDPred2 scores
ldpred2_path <- file.path(prs_out_dir, "single_ldpred2", "prs_scores.tsv")
if (file.exists(ldpred2_path)) {
  ldpred2_scores <- read_tsv(ldpred2_path)
  glimpse(ldpred2_scores)
}
```

---

# Section 3: Visualize PRS Performance

```{r visualize-prs}
# 1. Compare R² across single-ancestry methods
if (exists("perf")) {
  ggplot(perf, aes(x = reorder(method, R2), y = R2, fill = method)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = round(R2, 3)), vjust = -0.3, size = 3) +
    labs(
      title = "PRS Performance (R²) Across Single-Ancestry Methods",
      x = "Method",
      y = "Variance Explained (R²)",
      fill = "Method"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# 2. Compare AUC across methods (for binary traits)
if (exists("perf") && any(!is.na(perf$AUC))) {
  ggplot(perf, aes(x = reorder(method, AUC), y = AUC, fill = method)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = round(AUC, 3)), vjust = -0.3, size = 3) +
    labs(
      title = "PRS Performance (AUC) Across Single-Ancestry Methods",
      x = "Method",
      y = "Area Under ROC Curve (AUC)",
      fill = "Method"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# 3. PRS score distribution (example with PRSice2)
if (exists("prs_scores")) {
  ggplot(prs_scores, aes(x = PRS)) +
    geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
    labs(
      title = "PRS Score Distribution (PRSice2)",
      x = "Polygenic Risk Score",
      y = "Number of Samples"
    ) +
    theme_minimal()
}
```

---

# Section 4: Interactive Analysis

```{r interactive-analysis}
# Identify best-performing method
if (exists("perf")) {
  best_method <- perf %>%
    arrange(desc(R2)) %>%
    slice(1)
  print(best_method)
}

# Compare p-values across methods
if (exists("perf")) {
  perf <- perf %>%
    mutate(significant = p_value < 0.05)
  sig_counts <- perf %>%
    count(significant)
  print(sig_counts)
}

# Correlate PRS scores with phenotype (if phenotype data available)
if (exists("prs_scores")) {
  pheno_path <- file.path(PRS_OUT_DIR, "target_phenotype.tsv")
  if (file.exists(pheno_path)) {
    pheno <- read_tsv(pheno_path, col_names = c("IID", "pheno"))
    combined <- prs_scores %>%
      left_join(pheno, by = "IID")
    cor_test <- cor.test(combined$PRS, combined$pheno)
    print(cor_test)
  }
}
```

---

# Next Steps

- Compare single-ancestry PRS to multi-ancestry PRS (see multi-ancestry PRS tutorial)
- Validate PRS in independent holdout samples
- Export top-performing PRS scores for downstream analysis