---
title: "Lab: Visualizing and Working with SNP Heritability Outputs"
author: "GDCGenomicsQC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab: Visualizing and Working with SNP Heritability Outputs (R + Tidyverse)

This lab covers interacting with output files from the **SNP Heritability Estimation Pipeline** using R and tidyverse. You will work with per-ancestry heritability estimates, standard errors, and compare methods.

**Estimated time**: 45 minutes

**Prerequisites**:
- Completed the heritability tutorial (outputs in `OUT_DIR/03-snpHeritability/`)
- R (≥4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

---

# Setup: Load Libraries and Set Paths

```{r load-libraries}
library(tidyverse)
OUT_DIR <- "/path/to/your/pipeline/output"
herit_dir <- file.path(OUT_DIR, "03-snpHeritability")
```

---

# Section 1: Heritability Output Files Reference

| File Path | Description |
|----------|------------|
| `{subset}/03-snpHeritability/mash_output.csv` | Per-ancestry heritability estimates (from MASH) |
| `{subset}/03-snpHeritability/mash_config.json` | Configuration used for MASH run |
| `{subset}/03-snpHeritability/mash.log` | MASH execution log |

---

# Section 2: Load and Inspect Heritability Data

```{r load-data}
# Load heritability estimates (MASH output - CSV format with header)
subset <- "AFR"  # or EUR, etc.
herit_path <- file.path(herit_dir, subset, "03-snpHeritability", "mash_output.csv")
if (file.exists(herit_path)) {
  herit <- read_csv(herit_path)
  glimpse(herit)
}

# Load configuration (for reference)
config_path <- file.path(herit_dir, subset, "03-snpHeritability", "mash_config.json")
if (file.exists(config_path)) {
  comp <- read_tsv(comp_path)
  glimpse(comp)
}

# Calculate 95% confidence intervals
if (exists("herit")) {
  herit <- herit %>%
    mutate(
      lower_ci = h2 - 1.96 * SE,
      upper_ci = h2 + 1.96 * SE
    )
  print(herit)
}
```

---

# Section 3: Visualize Heritability Estimates

```{r visualize-heritability}
# 1. Forest plot of heritability per ancestry
if (exists("herit")) {
  ggplot(herit, aes(x = reorder(Ancestry, h2), y = h2, color = method)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    labs(
      title = "SNP Heritability Estimates per Ancestry",
      x = "Ancestry",
      y = "Heritability (h²)",
      color = "Method"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# 2. Compare heritability across methods
if (exists("comp")) {
  ggplot(comp, aes(x = method, y = h2, fill = Ancestry)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_errorbar(aes(ymin = h2 - 1.96*SE, ymax = h2 + 1.96*SE), 
                  position = position_dodge(0.9), width = 0.2) +
    labs(
      title = "Heritability Comparison Across Methods",
      x = "Estimation Method",
      y = "Heritability (h²)",
      fill = "Ancestry"
    ) +
    theme_minimal()
}

# 3. Standard error by sample size (if sample size data available)
if (exists("herit") && "n_samples" %in% colnames(herit)) {
  ggplot(herit, aes(x = n_samples, y = SE, color = method)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = "Standard Error vs Sample Size",
      x = "Number of Samples",
      y = "Standard Error (SE)",
      color = "Method"
    ) +
    theme_minimal()
}
```

---

# Section 4: Interactive Analysis

```{r interactive-analysis}
# Identify ancestries with significant heritability (p < 0.05)
if (exists("herit")) {
  sig_herit <- herit %>%
    filter(p_value < 0.05) %>%
    arrange(p_value)
  print(sig_herit)
}

# Compare h² between EUR and AFR
if (exists("herit")) {
  eur_afr <- herit %>%
    filter(Ancestry %in% c("EUR", "AFR")) %>%
    select(Ancestry, h2, SE)
  print(eur_afr)
}

# Calculate proportion of significant estimates
if (exists("herit")) {
  sig_prop <- herit %>%
    group_by(method) %>%
    summarise(prop_sig = mean(p_value < 0.05), .groups = "drop")
  print(sig_prop)
}
```

---

# Next Steps

- Compare heritability estimates across different phenotypes
- Test sensitivity to PC adjustment (vary npc parameter)
- Proceed to the single-ancestry PRS visualization lab for PRS outputs