---
title: "Lab: Visualizing Local Ancestry Outputs"
author: "Christian Coffman"
date: "2026-06-12"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

Local ancestry deconvolves the two haplotypes at each genomic position into their ancestral origins. The pipeline phases with ShapeIt4 then runs RFMix to infer ancestry tracts along each chromosome. This is the most computationally intensive step (phasing is the bottleneck), but the outputs are intuitive: ancestry blocks along each chromosome, per-SNP posterior probabilities, and genome-wide ancestry proportions.

## Input

```yaml
# sandboxToy.yaml lines 66-71
localAncestry:
    RFMIX: false
    chromosomes: [20, 21]
```

When `RFMIX: true`, the pipeline runs phasing + RFMix on the specified chromosomes (use all autosomes for production). See the [Multi-Ancestry PRS Tutorial](../tutorial_prs_multi.rst) for the full local ancestry pipeline reference.

## Output

### Global Ancestry Proportions

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy_chr"
local_anc_dir <- file.path(OUT_DIR, "02-localAncestry")

anc_full <- read_tsv(file.path(local_anc_dir, "ancestry_full.txt"))
glimpse(anc_full)

high_afr <- anc_full |> filter(AFR > 0.2) |> arrange(desc(AFR))
print(high_afr)
```

### Ancestry Tracts (MSP)

```{r, message=FALSE, warning=FALSE, error=FALSE}
chr <- 20
msp <- read_tsv(file.path(local_anc_dir, paste0("chr", chr, ".lai.msp.tsv")))
glimpse(msp)

sample_for_tract <- colnames(msp)[grep("^[^.]+\\.\\d+$", colnames(msp))][1:10]

msp |>
  select(chm, spos, epos, all_of(sample_for_tract)) |>
  pivot_longer(cols = all_of(sample_for_tract), names_to = "sample_hap", values_to = "ancestry_code") |>
  mutate(bp_mb = spos / 1e6) |>
  ggplot(aes(x = bp_mb, xend = epos / 1e6, y = sample_hap, yend = sample_hap, color = as.factor(ancestry_code))) +
  geom_segment(linewidth = 3) +
  labs(x = "Position (Mb)", y = "Sample (Haplotype)", color = "Ancestry",
       title = paste0("Ancestry Tracts on Chromosome ", chr)) +
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

### Per-SNP Local Ancestry

```{r, message=FALSE, warning=FALSE, error=FALSE}
snp_la <- read_tsv(file.path(local_anc_dir, "per_snp_local_ancestry.tsv"),
  col_names = c("SNP", "CHR", "BP", "IID", "predicted_ancestry",
                "prob_AFR", "prob_EUR", "prob_AMR"))
glimpse(snp_la)

snp_la |>
  ggplot(aes(x = BP, y = prob_EUR, color = as.factor(CHR))) +
  geom_point(size = 0.5, alpha = 0.7) +
  facet_wrap(~CHR, scales = "free_x", nrow = 2) +
  labs(x = "Base Pair Position", y = "Posterior Probability (EUR)",
       color = "Chromosome",
       title = "Per-SNP European Local Ancestry Probability") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Discussion

1. **Samples with many short ancestry tracts — is this real admixture or a phasing artifact?** How would you distinguish them?
2. **If most samples are assigned a single ancestry**, does that mean RFMix failed or that the population is genuinely unadmixed? Check the reference panel composition.
3. **How do the global ancestry proportions from `ancestry_full.txt` compare to the RFMix posterior probabilities from the classification step?** (See lab03.)

See the [Multi-Ancestry PRS Tutorial](../tutorial_prs_multi.rst) for troubleshooting, parameter tuning (generations, EM iterations), and comparing RFMix to other local ancestry methods.
