---
title: "Lab: Visualizing Local Ancestry Outputs"
author: "Christian Coffman"
date: "2026-06-12"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

Local ancestry deconvolves the two haplotypes at each genomic position into their ancestral origins. The pipeline phases with ShapeIt4 then runs RFMix to infer ancestry tracts along each chromosome. This is the most computationally intensive step (phasing is the bottleneck), but the outputs are intuitive: ancestry blocks along each chromosome, per-SNP posterior probabilities, and genome-wide ancestry proportions.

## Input

```yaml
# sandboxToy.yaml lines 66-71 (relevant keys)
localAncestry:
    RFMIX: false          # Toggle local ancestry analysis: false = skip, true = run ShapeIt4 phasing + RFMix
    chromosomes: [20, 21] # Which chromosomes to phase and analyze; use pipeline-wide list or custom subset
```

When `RFMIX: true`, the pipeline runs phasing + RFMix on the specified chromosomes (use all autosomes for production). See the [Multi-Ancestry PRS Tutorial](../tutorial_prs_multi.rst) for the full local ancestry pipeline reference.

## Output

### Global Ancestry Proportions

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
# CHANGE THIS to match your OUT_DIR from the config YAML
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy_chr"
local_anc_dir <- file.path(OUT_DIR, "02-localAncestry")

anc_full <- read_tsv(file.path(local_anc_dir, "ancestry_full.txt"))
glimpse(anc_full)

high_afr <- anc_full |> filter(AFR > 0.2) |> arrange(desc(AFR))
print(high_afr)
```

### Ancestry Tracts (MSP)

```{r, message=FALSE, warning=FALSE, error=FALSE}
chr <- 20
msp <- read_tsv(file.path(local_anc_dir, paste0("chr", chr, ".lai.msp.tsv")))
glimpse(msp)

sample_for_tract <- colnames(msp)[grep("^[^.]+\\.\\d+$", colnames(msp))][1:10]

msp |>
  select(chm, spos, epos, all_of(sample_for_tract)) |>
  pivot_longer(cols = all_of(sample_for_tract), names_to = "sample_hap", values_to = "ancestry_code") |>
  mutate(bp_mb = spos / 1e6) |>
  ggplot(aes(x = bp_mb, xend = epos / 1e6, y = sample_hap, yend = sample_hap, color = as.factor(ancestry_code))) +
  geom_segment(linewidth = 3) +
  labs(x = "Position (Mb)", y = "Sample (Haplotype)", color = "Ancestry",
       title = paste0("Ancestry Tracts on Chromosome ", chr)) +
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

### Per-SNP Local Ancestry

```{r, message=FALSE, warning=FALSE, error=FALSE}
snp_la <- read_tsv(file.path(local_anc_dir, "per_snp_local_ancestry.tsv"),
  col_names = c("SNP", "CHR", "BP", "IID", "predicted_ancestry",
                "prob_AFR", "prob_EUR", "prob_AMR"))
glimpse(snp_la)

snp_la |>
  ggplot(aes(x = BP, y = prob_EUR, color = as.factor(CHR))) +
  geom_point(size = 0.5, alpha = 0.7) +
  facet_wrap(~CHR, scales = "free_x", nrow = 2) +
  labs(x = "Base Pair Position", y = "Posterior Probability (EUR)",
       color = "Chromosome",
       title = "Per-SNP European Local Ancestry Probability") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Discussion

1. **Samples with many short ancestry tracts — is this real admixture or a phasing artifact?** How would you distinguish them?
2. **If most samples are assigned a single ancestry**, does that mean RFMix failed or that the population is genuinely unadmixed? Check the reference panel composition.
3. **How do the global ancestry proportions from `ancestry_full.txt` compare to the RFMix posterior probabilities from the classification step?** (See the ancestry lab.)
4. **How does local ancestry inform heritability analysis?** If a trait-associated variant lies in an AFR-derived haplotype block, its effect size may differ from the same variant in an EUR-derived block. How would you model ancestry-specific effect sizes in a heritability or PRS framework?
5. **What happens to phasing accuracy in recently admixed populations?** ShapeIt4 uses a reference panel for phasing — if the panel is mostly EUR but your sample is AFR/EUR admixed, which haplotypes will be phased more accurately? How could phasing errors propagate into RFMix ancestry calls?
6. **The downstream PRS lab uses global ancestry-adjusted scores. How would ancestry-aware PRS differ if you replaced global ancestry with local ancestry at each variant?** Would you expect better prediction in admixed individuals? What additional data would you need?
7. **If you have trio data (parent-offspring) with local ancestry, how could you distinguish recent admixture (within the last 2–3 generations) from older admixture?** (Hint: tract length distribution is informative — longer tracts mean fewer generations since admixture.)

See the [Multi-Ancestry PRS Tutorial](../tutorial_prs_multi.rst) for troubleshooting, parameter tuning (generations, EM iterations), and comparing RFMix to other local ancestry methods.
