---
title: "Lab: Visualizing and Working with Global Ancestry Outputs"
author: "GDCGenomicsQC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab: Visualizing and Working with Global Ancestry Outputs (R + Tidyverse)

This lab covers interacting with output files from the **Global Ancestry Classification Pipeline** using R and tidyverse. You will work with posterior probabilities, ancestry classifications, and confusion matrices.

**Estimated time**: 45 minutes

**Prerequisites**:
- Completed the ancestry classification tutorial (outputs in `OUT_DIR/01-globalAncestry/`)
- R (≥4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

---

# Setup: Load Libraries and Set Paths

```{r load-libraries}
library(tidyverse)
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy"
global_anc_dir <- file.path(OUT_DIR, "01-globalAncestry")
```

---

# Section 1: Global Ancestry Output Files Reference

| File Path | Description |
|----------|------------|
| `01-globalAncestry/posterior_probabilities.tsv` | Per-sample posterior probabilities for each ancestry |
| `01-globalAncestry/sample_coords.tsv` | Sample coordinates in PCA/UMAP space |
| `01-globalAncestry/ref_coords.tsv` | Reference coordinates in PCA/UMAP space |
| `01-globalAncestry/ancestry_classifications.tsv` | Predicted ancestry and confidence per sample |
| `01-globalAncestry/keep_{ANC}.txt` | Sample IDs passing ancestry threshold for each ancestry |
| `01-globalAncestry/posterior_probability_stacked_{model}.svg` | Pre-rendered stacked posterior probability plot (model = pca/umap) |

---

# Section 2: Load and Inspect Global Ancestry Data

```{r load-data}
# Load posterior probabilities (PCA model example)
post_prob_path <- file.path(global_anc_dir, "posterior_probabilities.tsv")
if (file.exists(post_prob_path)) {
  post_prob <- read_tsv(post_prob_path)  # Columns: IID, pca_AFR, pca_AMR, pca_EUR, pca_SAS
  glimpse(post_prob)
}

# Load ancestry classifications
class_path <- file.path(global_anc_dir, "ancestry_classifications.tsv")
if (file.exists(class_path)) {
  class <- read_tsv(class_path)  # Columns: IID, pca_predicted, pca_confidence
  glimpse(class)
}

# Join and pivot datasets for visualization
if (exists("post_prob") && exists("class")) {
  anc_data <- post_prob %>%
    left_join(class, by = "IID") %>%
    mutate(ancestry = str_sub(IID, 1, 3))
  glimpse(anc_data)
}

table(anc_data$ancestry, anc_data$umap_predicted)
table(anc_data$ancestry, anc_data$pca_predicted)
```

---

# Section 3: Visualize Global Ancestry

```{r visualize-ancestry}
anc_data |>
    pivot_longer(
      cols = pca_EUR:pca_AFR,
      names_to = "supposed_ancestry",
      values_to = "posterior_prob"
    ) |>
  slice_head(n = 50, by = ancestry) |>
  ggplot(aes(x = reorder(IID, posterior_prob), y = posterior_prob, fill = supposed_ancestry)) +
  geom_col(width = 1) +
  labs(
    title = "Per-Sample Posterior Probabilities for Global Ancestry (PCA)",
    x = "Sample ID",
    y = "Posterior Probability",
    fill = "Ancestry"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~ancestry, scales = "free_x")
```


# Next Steps

- Compare PCA vs UMAP posterior probability distributions
- Analyze discordance between predicted ancestry and reported race
- Proceed to the heritability visualization lab for heritability outputs