---
title: "Lab: Visualizing and Working with Global Ancestry Outputs"
---

Interact with output files from the Global Ancestry Classification Pipeline.
Work with posterior probabilities, ancestry classifications, and confusion matrices.

**Prerequisites:**
- Completed ancestry classification tutorial (outputs in `OUT_DIR/01-globalAncestry/`)
- R (>= 4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

```{r setup}
conda_env <- '/scratch.global/R25_files/r25_r_env/lib/R/library'
.libPaths(unique(c(.Library.site, .Library)))
.libPaths(c(conda_env, .libPaths()))
library(tidyverse)

OUT_DIR <- "/scratch.global/GDC/r25outputs/toy"
global_anc_dir <- file.path(OUT_DIR, "01-globalAncestry")
```

## Load posterior probabilities

```{r}
post_prob_path <- file.path(global_anc_dir, "posterior_probabilities.tsv")
if (file.exists(post_prob_path)) {
  post_prob <- read_tsv(post_prob_path)
  glimpse(post_prob)
}
```

## Load ancestry classifications

```{r}
class_path <- file.path(global_anc_dir, "ancestry_classifications.tsv")
if (file.exists(class_path)) {
  class <- read_tsv(class_path)
  glimpse(class)
}
```

## Join and pivot datasets for visualization

```{r}
if (exists("post_prob") && exists("class")) {
  anc_data <- post_prob %>%
    left_join(class, by = "IID") %>%
    mutate(ancestry = str_sub(IID, 1, 3))
  glimpse(anc_data)
}
```

## Confusion matrices

```{r}
table(anc_data$ancestry, anc_data$umap_predicted)
table(anc_data$ancestry, anc_data$pca_predicted)
```

## Visualize: per-sample posterior probabilities

```{r}
anc_data %>%
  pivot_longer(
    cols = pca_EUR:pca_AFR,
    names_to = "supposed_ancestry",
    values_to = "posterior_prob"
  ) %>%
  slice_head(n = 50, by = ancestry) %>%
  ggplot(aes(x = reorder(IID, posterior_prob), y = posterior_prob, fill = supposed_ancestry)) +
  geom_col(width = 1) +
  labs(
    title = "Per-Sample Posterior Probabilities for Global Ancestry (PCA)",
    x = "Sample ID",
    y = "Posterior Probability",
    fill = "Ancestry"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank()) +
  facet_wrap(~ancestry, scales = "free_x")
```
