---
title: "Lab: Visualizing and Working with SNP Heritability Outputs"
---

Interact with output files from the SNP Heritability Estimation Pipeline.
Work with per-ancestry heritability estimates, standard errors, and compare methods.

**Prerequisites:**
- Completed heritability tutorial (outputs in `OUT_DIR/03-snpHeritability/`)
- R (>= 4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

```{r setup}
conda_env <- '/scratch.global/R25_files/r25_r_env/lib/R/library'
.libPaths(unique(c(.Library.site, .Library)))
.libPaths(c(conda_env, .libPaths()))
library(tidyverse)
OUT_DIR <- "/scratch.global/GDC/r25outputs/toy"
herit_dir <- file.path(OUT_DIR, "03-snpHeritability")
```

## Load heritability estimates

```{r}
subset <- "AFR"
herit_path <- file.path(herit_dir, subset, "03-snpHeritability", "mash_output.csv")
if (file.exists(herit_path)) {
  herit <- read_csv(herit_path)
  glimpse(herit)
}
```

## Calculate 95% confidence intervals

```{r}
if (exists("herit")) {
  herit <- herit %>%
    mutate(
      lower_ci = h2 - 1.96 * SE,
      upper_ci = h2 + 1.96 * SE
    )
  print(herit)
}
```

## Forest plot of heritability per ancestry

```{r}
if (exists("herit")) {
  ggplot(herit, aes(x = reorder(Ancestry, h2), y = h2, color = method)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    labs(
      title = "SNP Heritability Estimates per Ancestry",
      x = "Ancestry",
      y = expression("Heritability (h"^2*")"),
      color = "Method"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Identify ancestries with significant heritability

```{r}
if (exists("herit")) {
  sig_herit <- herit %>%
    filter(p_value < 0.05) %>%
    arrange(p_value)
  print(sig_herit)
}
```
