---
title: "Lab: Visualizing and Working with Local Ancestry Outputs"
---

Interact with output files from the Local Ancestry Pipeline (RFMix).
Work with local ancestry tracts, per-SNP ancestry calls, and posterior probabilities.

**Prerequisites:**
- Pipeline run with `localAncestry: RFMIX: true` in config
- Local ancestry outputs present in `OUT_DIR/02-localAncestry/`
- R (>= 4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

```{r setup}
conda_env <- '/scratch.global/R25_files/r25_r_env/lib/R/library'
.libPaths(unique(c(.Library.site, .Library)))
.libPaths(c(conda_env, .libPaths()))
library(tidyverse)

OUT_DIR <- "/path/to/your/pipeline/output"
local_anc_dir <- file.path(OUT_DIR, "02-localAncestry")
chr <- 20
```

## Load RFMix global ancestry

```{r}
anc_full_path <- file.path(local_anc_dir, "ancestry_full.txt")
if (file.exists(anc_full_path)) {
  anc_full <- read_tsv(anc_full_path)
  glimpse(anc_full)
}
```

## Load per-chromosome MSP file

```{r}
msp_path <- file.path(local_anc_dir, paste0("chr", chr, ".lai.msp.tsv"))
if (file.exists(msp_path)) {
  msp <- read_tsv(msp_path)
  glimpse(msp)
}
```

## Load per-chromosome fragment file (ancestry tracts)

```{r}
fb_path <- file.path(local_anc_dir, paste0("chr", chr, ".lai.fb.tsv"))
if (file.exists(fb_path)) {
  fb <- read_tsv(fb_path)
  glimpse(fb)
}
```

## Load per-SNP local ancestry

```{r}
snp_la_path <- file.path(local_anc_dir, "per_snp_local_ancestry.tsv")
if (file.exists(snp_la_path)) {
  snp_la <- read_tsv(snp_la_path,
    col_names = c("SNP", "CHR", "BP", "IID", "predicted_ancestry",
                  "prob_AFR", "prob_EUR", "prob_AMR"))
  glimpse(snp_la)
}
```

## Manhattan plot of local ancestry

```{r}
if (exists("snp_la")) {
  ggplot(snp_la, aes(x = BP, y = prob_EUR, color = as.factor(CHR))) +
    geom_point(size = 0.5, alpha = 0.7) +
    facet_wrap(~CHR, scales = "free_x", nrow = 2) +
    labs(
      title = "Per-SNP European Local Ancestry Probability",
      x = "Base Pair Position",
      y = "Posterior Probability (EUR)",
      color = "Chromosome"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Identify samples with high African ancestry proportion

```{r}
if (exists("anc_full")) {
  high_afr <- anc_full %>%
    filter(AFR > 0.2) %>%
    arrange(desc(AFR))
  print(high_afr)
}
```
