---
title: "Lab: Visualizing Local Ancestry Outputs"
---

## Overview

Local ancestry deconvolves the two haplotypes at each genomic position into their ancestral origins. The pipeline uses RFMix (via the `phase.smk` and `ancestry.smk` rules) to phase genotypes and infer ancestry tracts. The key outputs include global ancestry proportions (`ancestry_full.txt`), per-chromosome ancestry tracts in MSP format, and per-SNP local ancestry calls.

The genotypes are **synthetic** — resampled from 1000 Genomes haplotypes and available on Dataverse. Because the underlying haplotypes have clear population labels, local ancestry assignments should be **highly confident**, with long contiguous tracts and few switch errors.

## Goals & Expectations

Frame your exploration around these questions:

1. **Global ancestry proportions** — Do the RFMix-derived global proportions match the per-sample ancestry composition expected from the known population structure?
2. **Local ancestry tracts** — Are tracts long and contiguous, or are there many short fragments suggesting phasing uncertainty?
3. **Per-SNP posterior probabilities** — Do most SNPs have posterior probability > 0.9 for a single ancestry?
4. **High-proportion samples** — Which samples have elevated African ancestry (>20%)? Does this match their known label?

Given the **synthetic origin** and the clean 1KG reference, you should expect **long ancestry blocks** and **high-confidence calls** with minimal uncertainty.

## Inputs & Outputs

- **Inputs:** Phased study PGEN/VCF (from ShapeIt4); 1000 Genomes phased reference panel; RFMix configuration (number of generations, EM iterations).
- **Outputs:** `ancestry_full.txt` (genome-wide ancestry proportions per sample), `chr{chr}.lai.msp.tsv` (per-chromosome ancestry tracts), `chr{chr}.lai.fb.tsv` (forwards-backwards probabilities), `per_snp_local_ancestry.tsv` (combined SNP-level calls).
- **Pipeline direction:** The global ancestry proportions (`ancestry_full.txt`) feed back into the global ancestry classification step (as the `rfmix_` posterior probabilities). The per-SNP calls can be used for downstream local-ancestry-aware analyses.

## Upstream Pipeline Steps

Local ancestry inference is one of the most computationally intensive steps. Key upstream dependencies:

1. **Initial QC + Standard QC** — RFMix requires high-quality variants. Low-quality variants create false-positive ancestry switches.
2. **Unrelated sample extraction** (`extractUnrelatedSubjects` in `phase.smk`) — Removes related samples before phasing and RFMix. **Related samples share long haplotypes by descent**, which would be misidentified as shared ancestry tracts.
3. **Phasing with ShapeIt4** (`phaseShapeIt4` in `phase.smk`) — RFMix requires **phased haplotypes** as input (not unphased genotypes). ShapeIt4 uses a reference panel and statistical models to infer the most likely haplotype phase. Phase errors propagate directly to ancestry-switch errors.
4. **Crossmap** — Ensures study variants and reference variants use the same genome build and reference allele.
5. **Per-chromosome processing** — Both phasing and RFMix run per chromosome to avoid memory exhaustion on large datasets.

**Why independence matters here:** Phasing assumes unrelated individuals — related samples share long IBD segments that violate the phasing model's independence assumptions. RFMix then compares each study haplotype against reference haplotypes; related study samples would produce artificially long shared tracts.

## Before You Begin

- **Required files:** `ancestry_full.txt`, `chr{chr}.lai.msp.tsv`, `chr{chr}.lai.fb.tsv`, `per_snp_local_ancestry.tsv` in `{local_anc_dir}/`
- **Disk space:** MSP files can be several GB per chromosome (11K × 377M variants); the `per_snp_local_ancestry.tsv` is a summarized subset
- **Expected render time:** 30–60 seconds for MSP loading; < 5 seconds for aggregated files
- **Known issue:** The forward-backward `.fb.tsv` files can be very large — if loading fails, try reading only the first chromosome or using `data.table::fread` with `nrows`
- **Reference:** RFMix was run with 2 EM steps and the default number of generations (12). The reference panel is 1000 Genomes Phase 3.

## Setup

```{r setup}
conda_env <- '/scratch.global/R25_files/r25_r_env/lib/R/library'
.libPaths(unique(c(.Library.site, .Library)))
.libPaths(c(conda_env, .libPaths()))
library(tidyverse)

OUT_DIR <- "/path/to/your/pipeline/output"
local_anc_dir <- file.path(OUT_DIR, "02-localAncestry")
chr <- 20
```

## Global Ancestry Proportions

**File:** `{local_anc_dir}/ancestry_full.txt`

**Format:** TSV with one row per sample. Columns include `IID` followed by one column per ancestry (e.g., `AFR`, `EUR`, `AMR`). Each value is the genome-wide proportion of that ancestry averaged across both haplotypes.

**Pipeline step:** Generated by `workflow/scripts/rfmix_to_ancestry.R` (consolidated from gai.R) in the `run_rfmix` aggregate rule (`ancestry.smk`). It reads all per-chromosome `.msp.tsv` files, computes segment-length-weighted ancestry proportions, and writes the global table.

```{r}
anc_full_path <- file.path(local_anc_dir, "ancestry_full.txt")
if (file.exists(anc_full_path)) {
  anc_full <- read_tsv(anc_full_path)
  glimpse(anc_full)
}
```

## Per-Chromosome MSP File

**File:** `{local_anc_dir}/chr{chr}.lai.msp.tsv`

**Format:** RFMix "most recent shared ancestor" output. Each row is a genomic segment; columns include `chm`, `spos`, `epos`, `snps`, and one column per haplotype per sample with the ancestry integer code.

**Pipeline step:** Generated by RFMix (via `shapeit4` + `rfmix`) in the `run_rfmix` rule. The per-chromosome phasing is done by `phase.smk` (`phaseShapeit4`), followed by RFMix on the unrelated subset.

```{r}
msp_path <- file.path(local_anc_dir, paste0("chr", chr, ".lai.msp.tsv"))
if (file.exists(msp_path)) {
  msp <- read_tsv(msp_path)
  glimpse(msp)
}
```

## Fragment File (Ancestry Tracts)

**File:** `{local_anc_dir}/chr{chr}.lai.fb.tsv`

**Format:** Forwards-backwards posterior probabilities from RFMix. One row per SNP per sample, with columns for the probability of each ancestry at that SNP.

**Pipeline step:** Generated alongside the MSP file by RFMix. Used internally by `rfmix_to_ancestry.R` to compute per-SNP ancestry calls.

```{r}
fb_path <- file.path(local_anc_dir, paste0("chr", chr, ".lai.fb.tsv"))
if (file.exists(fb_path)) {
  fb <- read_tsv(fb_path)
  glimpse(fb)
}
```

## Per-SNP Local Ancestry

**File:** `{local_anc_dir}/per_snp_local_ancestry.tsv`

**Format:** Combined output across all chromosomes. Columns: `SNP`, `CHR`, `BP`, `IID`, `predicted_ancestry`, `prob_AFR`, `prob_EUR`, `prob_AMR`.

**Pipeline step:** Generated by a post-processing step in the `run_rfmix` rule that collapses the fragment files into a single per-SNP calls table.

```{r}
snp_la_path <- file.path(local_anc_dir, "per_snp_local_ancestry.tsv")
if (file.exists(snp_la_path)) {
  snp_la <- read_tsv(snp_la_path,
    col_names = c("SNP", "CHR", "BP", "IID", "predicted_ancestry",
                  "prob_AFR", "prob_EUR", "prob_AMR"))
  glimpse(snp_la)
}
```

## Manhattan Plot of Local Ancestry

Visualize per-SNP posterior probability for European ancestry across the genome. Each point is a SNP; color groups chromosomes.

```{r}
if (exists("snp_la")) {
  ggplot(snp_la, aes(x = BP, y = prob_EUR, color = as.factor(CHR))) +
    geom_point(size = 0.5, alpha = 0.7) +
    facet_wrap(~CHR, scales = "free_x", nrow = 2) +
    labs(
      title = "Per-SNP European Local Ancestry Probability",
      x = "Base Pair Position",
      y = "Posterior Probability (EUR)",
      color = "Chromosome"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

**Expectation:** For a sample of predominantly European ancestry, you should see long stretches with prob_EUR near 1.0. Any dips to intermediate values may indicate admixture or phasing uncertainty. With synthetic data, tracts should be clean and probabilities extreme.

## Samples with High African Ancestry

Which samples have >20% African ancestry according to RFMix?

```{r}
if (exists("anc_full")) {
  high_afr <- anc_full %>%
    filter(AFR > 0.2) %>%
    arrange(desc(AFR))
  print(high_afr)
}
```

**Expectation:** Only samples that are genuinely admixed or from African-ancestry populations should appear here. The synthetic resampling preserves the original 1KG population labels, so these proportions should match the known composition.

## Ancestry Tract Track Plot

Visualize ancestry tracts along a chromosome as colored segments. This karyogram-style plot shows ancestry switches most intuitively.

```{r}
if (exists("msp")) {
  sample_for_tract <- colnames(msp)[grep("^[^.]+\\.\\d+$", colnames(msp))][1:10]

  msp_long <- msp %>%
    select(chm, spos, epos, all_of(sample_for_tract)) %>%
    pivot_longer(cols = all_of(sample_for_tract), names_to = "sample_hap", values_to = "ancestry_code")

  msp_long %>%
    mutate(bp_mb = spos / 1e6) %>%
    ggplot(aes(x = bp_mb, xend = epos / 1e6, y = sample_hap, yend = sample_hap, color = as.factor(ancestry_code))) +
    geom_segment(linewidth = 3) +
    labs(
      title = paste0("Ancestry Tracts on Chromosome ", chr),
      x = "Position (Mb)",
      y = "Sample (Haplotype)",
      color = "Ancestry"
    ) +
    theme_minimal() +
    theme(axis.text.y = element_blank())
}
```

**Expectation:** Each sample-haplotype row should show long, contiguous blocks of a single color. Many short alternating blocks suggest phasing errors or genuine admixture. With synthetic 1KG data, tracts should be long and switches rare.

## Local vs. Global Concordance

Compare the RFMix-derived genome-wide ancestry proportions (from `ancestry_full.txt`) with the random forest `rfmix_*` classification probabilities. They should agree closely.

```{r}
if (exists("anc_full") && exists("class_prob")) {
  local_global <- anc_full %>%
    inner_join(class_prob, by = "IID")

  p_afr <- local_global %>%
    ggplot(aes(x = AFR, y = rfmix_AFR)) +
    geom_point(alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    labs(
      title = "RFMix Global AFR Proportion vs. Random Forest AFR Probability",
      x = "RFMix Global Proportion (from ancestry_full.txt)",
      y = "Random Forest Probability (from classificationProbabilities.tsv)"
    ) +
    theme_minimal()
  print(p_afr)
}
```

**Expectation:** Points should cluster near the y = x line. Systematic deviation suggests one of the two methods is biased. With clean synthetic data, agreement should be high (R² > 0.95).

## Check Your Understanding

1. **Refix a sample with multiple ancestry switches on chr20.** Is this biologically real (recent admixture) or an artifact (phasing error)? How would you distinguish them?
2. **Which is more robust to phasing errors: the segment-length-weighted global ancestry from `.msp.tsv` or the Q-mean from `.rfmix.Q` files?** Why?
3. **If the local ancestry tracts show very short blocks (< 1 Mb) for most samples**, what does that suggest about the RFMix parameters or the reference panel?

## Troubleshooting

| Problem | Likely Cause | Solution |
|---------|-------------|----------|
| `.fb.tsv` is too large to load | FB files are chr × samples × SNPs | Use `data.table::fread` with `nrows = 10000`, or load only one chromosome |
| MSP file has no ancestry columns | RFMix failed silently | Check `run_rfmix` log; verify the phased VCF and reference panel are on the same build |
| Global proportions don't sum to 1 | Numerical precision in segment-length weights | The `rfmix_to_ancestry.R` script sums proportions; slight deviations (< 0.01) are expected |
| Tract plot shows only one color | All samples assigned a single ancestry | Check if RFMix received only one reference population; verify the `.ref.Classes.txt` file |

## Going Further

- **Recombination hotspot detection**: Count ancestry switches per Mb for each sample. Do switch rates correlate with known recombination hotspots (e.g., from HapMap)?
- **Per-SNP ancestry at known genes**: Extract local ancestry calls at a gene of interest (e.g., `LCT` on chr2). Do samples with European ancestry at that locus have a different phenotype distribution?
- **Compare RFMix to FLARE**: In the literature, FLARE and RFMix give similar local ancestry calls. If you have FLARE output, compute the per-SNP concordance rate between the two methods.
