---
title: "Lab: Visualizing and Working with Local Ancestry Outputs"
author: "GDCGenomicsQC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab: Visualizing and Working with Local Ancestry Outputs (R + Tidyverse)

This lab covers interacting with output files from the **Local Ancestry Pipeline** (RFMix) using R and tidyverse. You will work with local ancestry tracts, per-SNP ancestry calls, and posterior probabilities.

**Estimated time**: 1 hour

**Prerequisites**:
- Pipeline run with `localAncestry: RFMIX: true` in config
- Local ancestry outputs present in `OUT_DIR/02-localAncestry/`
- R (≥4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

---

# Setup: Load Libraries and Set Paths

```{r load-libraries}
library(tidyverse)
OUT_DIR <- "/path/to/your/pipeline/output"
local_anc_dir <- file.path(OUT_DIR, "02-localAncestry")
chr <- 20  # Example chromosome
```

---

# Section 1: Local Ancestry Output Files Reference

| File Path | Description |
|----------|------------|
| `02-localAncestry/ancestry_full.txt` | Global ancestry per sample (from RFMix) |
| `02-localAncestry/chr{CHR}.lai.msp.tsv` | Per-chromosome MSP (marginal success probability) files |
| `02-localAncestry/chr{CHR}.lai.fb.tsv` | Per-chromosome fragment files (ancestry tracts) |
| `02-localAncestry/` | Contains all per-chromosome ancestry outputs |

---

# Section 2: Load and Inspect Local Ancestry Data

```{r load-data}
# Load RFMix global ancestry (from MSP files)
anc_full_path <- file.path(local_anc_dir, "ancestry_full.txt")
if (file.exists(anc_full_path)) {
  anc_full <- read_tsv(anc_full_path)
  glimpse(anc_full)
}

# Load per-chromosome MSP file (marginal success probability)
msp_path <- file.path(local_anc_dir, paste0("chr", chr, ".lai.msp.tsv"))
if (file.exists(msp_path)) {
  msp <- read_tsv(msp_path)
  glimpse(msp)
}

# Load per-chromosome fragment file (ancestry tracts)
fb_path <- file.path(local_anc_dir, paste0("chr", chr, ".lai.fb.tsv"))
if (file.exists(fb_path)) {
  fb <- read_tsv(fb_path)
  glimpse(fb)
}

# Load per-SNP local ancestry
snp_la_path <- file.path(local_anc_dir, "per_snp_local_ancestry.tsv")
if (file.exists(snp_la_path)) {
  snp_la <- read_tsv(snp_la_path, 
                     col_names = c("SNP", "CHR", "BP", "IID", "predicted_ancestry", 
                                   "prob_AFR", "prob_EUR", "prob_AMR"))
  glimpse(snp_la)
}

# Summarize tracts per sample
if (exists("tracts")) {
  tract_summary <- tracts %>%
    group_by(IID, ancestry) %>%
    summarise(
      n_tracts = n(),
      total_bp = sum(end - start, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(IID, desc(total_bp))
  print(tract_summary)
}
```

---

# Section 3: Visualize Local Ancestry

```{r visualize-local-ancestry}
# 1. Manhattan plot of local ancestry (per-SNP posterior prob for EUR)
if (exists("snp_la")) {
  ggplot(snp_la, aes(x = BP, y = prob_EUR, color = as.factor(CHR))) +
    geom_point(size = 0.5, alpha = 0.7) +
    facet_wrap(~CHR, scales = "free_x", nrow = 2) +
    labs(
      title = "Per-SNP European Local Ancestry Probability",
      x = "Base Pair Position",
      y = "Posterior Probability (EUR)",
      color = "Chromosome"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# 2. Tract length distribution
if (exists("tracts")) {
  tracts <- tracts %>%
    mutate(tract_length = end - start)
  
  ggplot(tracts, aes(x = tract_length / 1e6, fill = ancestry)) +
    geom_histogram(bins = 30, alpha = 0.7, position = "dodge") +
    labs(
      title = "Local Ancestry Tract Length Distribution",
      x = "Tract Length (Mb)",
      y = "Number of Tracts",
      fill = "Ancestry"
    ) +
    theme_minimal()
}

# 3. Proportion of genome per ancestry per sample
if (exists("tracts")) {
  genome_prop <- tracts %>%
    group_by(IID) %>%
    mutate(total_genome = sum(tract_length, na.rm = TRUE)) %>%
    group_by(IID, ancestry) %>%
    summarise(prop = sum(tract_length, na.rm = TRUE) / first(total_genome), 
              .groups = "drop")
  
  ggplot(genome_prop, aes(x = reorder(IID, -prop), y = prop, fill = ancestry)) +
    geom_bar(stat = "identity") +
    labs(
      title = "Proportion of Genome by Local Ancestry per Sample",
      x = "Sample ID",
      y = "Proportion of Genome",
      fill = "Ancestry"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
}
```

---

# Section 4: Interactive Analysis

```{r interactive-analysis}
# Identify samples with high African ancestry proportion
if (exists("genome_prop")) {
  high_afr <- genome_prop %>%
    filter(ancestry == "AFR", prop > 0.2) %>%
    arrange(desc(prop))
  print(high_afr)
}

# Filter SNPs with uncertain local ancestry (max posterior < 0.8)
if (exists("snp_la")) {
  uncertain_snp <- snp_la %>%
    rowwise() %>%
    mutate(max_prob = max(c(prob_AFR, prob_EUR, prob_AMR), na.rm = TRUE)) %>%
    filter(max_prob < 0.8) %>%
    ungroup()
  print(uncertain_snp)
}
```

---

# Next Steps

- Compare local ancestry tracts across chromosomes
- Overlay local ancestry with GWAS significant hits
- Proceed to the global ancestry visualization lab for global ancestry outputs