---
title: "Lab: Visualizing SNP Heritability Outputs"
author: "Christian Coffman"
date: "2026-06-12"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    embed-resources: true
---

## Overview

SNP heritability (h²) estimates the proportion of phenotypic variance explained by genetic variants. The pipeline uses MASH (or GCTA) to estimate h² per ancestry using a GRM computed from LD-pruned, crossmapped variants.

The genotypes are **synthetic** — resampled from 1000 Genomes haplotypes and available on Dataverse. The phenotype is simulated with a **known heritability of 0.4**. This gives you a ground-truth target: the pipeline's estimates should cluster near h² ≈ 0.4 across methods and ancestries.

## Goals & Expectations

Frame your exploration around these questions:

1. **Heritability recovery** — Do the estimated h² values center around the true value of 0.4? How much do they vary across ancestries?
2. **Method agreement** — Do different estimation methods (e.g., MASH, GCTA-REML) give similar results?
3. **Statistical significance** — Which ancestry-specific estimates are significantly different from zero (p < 0.05)?
4. **Confidence intervals** — How wide are the 95% CIs? Do they capture 0.4?

Given the **simulated phenotype** with h² = 0.4, you should expect estimates near 0.4 with modest variation. If estimates are systematically biased upward or downward, that suggests an issue in the GRM construction or the estimation model.

## Inputs & Outputs

- **Inputs:** LD-pruned, crossmapped PGEN (`f1.b38.ldpruned.*`); phenotype file (simulated with h² = 0.4); covariate file (age, sex, PCs); GRM from GCTA.
- **Outputs:** `mash_output.csv` (per-ancestry h² estimates with SE and p-value).
- **Pipeline direction:** These estimates quantify how much of the trait variance is genetic. They inform power calculations for GWAS and help prioritize ancestries for PRS modeling.

## Upstream Pipeline Steps

Heritability estimation is sensitive to variant quality, LD, and sample structure. The pipeline runs these steps first:

1. **Initial QC** — Removes variants and samples with high missingness that would bias the GRM.
2. **Standard QC** (`filterStandard.sh` → `Standard_QC.smk`) — Applies MAF (>1%), HWE (p > 1e-6), and heterozygosity filters. **Low-MAF variants inflate heritability** because their effects are poorly estimated; HWE deviations suggest genotyping error rather than true genetic signal.
3. **LD pruning** (`concatLdPruned`) — The GRM computed by GCTA assumes **independent genetic markers**. Including correlated SNPs would double-count shared signal and inflate h² estimates.
4. **Crossmap** (`crossmapLdPrunedToB38`) — Ensures all variants map to the same genome build for consistent GRM computation.
5. **GRM computation** (`checkRelatednessExtractUnrelated` via `gcta64 --make-grm`) — Computes the genetic relationship matrix from the LD-pruned variants.
6. **Relatedness filtering** (`king` → `unrelated_ids.txt`) — Removes one individual from each pair with pi-hat > 0.0884. **Related samples inflate heritability** because they share both phenotype and genotype more than expected by chance.
7. **Phenotype simulation** (external / pre-pipeline) — A phenotype was simulated with **known h² = 0.4** and added to the synthetic genotype data. This provides a ground truth that real studies lack.

**Why independence matters here:** The GRM is a covariance matrix among individuals. If related individuals are included, the off-diagonal GRM elements are non-zero due to recent shared ancestry, not due to the genetic architecture of the trait. This biases h² estimates upward (confounds population structure with heritability).

## Configuration

Heritability estimation uses the `snpHerit` key in `config.yaml`. Here is a mock excerpt showing surrounding context:

```yaml
# config.yaml (mock excerpt — heritability-related keys only)
OUT_DIR: "/scratch/.../output"
REF: "/scratch/.../1000G_reference"

thresholds:
  maf: 0.01
  hwe: 1e-6

ancestry:
  model: pca

snpHerit:
  grm_prefix: /scratch/.../output/full/f1.b38.ldpruned.unrelated
  # Path to the GCTA GRM binary files (no extension).
  # Expected files: .grm.bin, .grm.N.bin, .grm.id
```

- **`snpHerit.grm_prefix`**: The GRM was built by GCTA `--make-grm` from LD-pruned, crossmapped variants using only unrelated samples. The pipeline concatenates per-chromosome LD-pruned files (`concatLdPruned`), crossmaps them to GRCh38, computes the GRM via `checkRelatednessExtractUnrelated`, then estimates h² per ancestry with MASH.
- The phenotype was **simulated externally** with known h² = 0.4. In a real study, it would come from clinical measurements.

The `mash_output.csv` below contains the per-ancestry h² estimates derived from this GRM.

## Before You Begin

- **Required files:** `mash_output.csv` in `{herit_dir}/{subset}/03-snpHeritability/`; GRM binary files for the heatmap section
- **Disk space:** < 1 MB for CSV; GRM files can be 100 MB–2 GB
- **Expected render time:** < 5 seconds for CSV; 10–30 seconds for GRM heatmap (depends on n_samples)
- **Known issue:** If `mash_output.csv` is empty, the MASH solver may have failed to converge. Check the `run_snpHerit` log.
- **Reference:** The phenotype was simulated externally with h² = 0.4 to provide a known ground truth.

## Setup

The code chunk below adds the conda R library path so that pre-installed packages are found without a local `install.packages()`. The environment at `/scratch.global/R25_files/r25_r_env/` is shared across users and contains all packages used in these labs:

- **tidyverse** (ggplot2, dplyr, tidyr, readr, purrr, stringr, forcats, tibble) — data manipulation and plotting
- **data.table** — fast file reading via `fread()`

```{r setup}
conda_env <- '/scratch.global/R25_files/r25_r_env/lib/R/library'
.libPaths(unique(c(.Library.site, .Library)))
.libPaths(c(conda_env, .libPaths()))
library(tidyverse)

OUT_DIR <- "/path/to/your/pipeline/output"
herit_dir <- file.path(OUT_DIR, "03-snpHeritability")
```

## Load Heritability Estimates

**File:** `{herit_dir}/{subset}/03-snpHeritability/mash_output.csv`

**Format:** CSV with columns for ancestry, method, h² estimate, standard error (SE), and p-value. One row per ancestry-method combination.

**Pipeline step:** Generated by the `run_snpHerit` rule in `Snakefile`. It builds a GRM (`checkRelatednessExtractUnrelated` → `concatLdPruned` → `crossmapLdPrunedToB38`), then runs MASH (or GCTA) to estimate variance components per ancestry.

```{r}
subset <- "AFR"
herit_path <- file.path(herit_dir, subset, "03-snpHeritability", "mash_output.csv")
if (file.exists(herit_path)) {
  herit <- read_csv(herit_path)
  glimpse(herit)
}
```

## Calculate 95% Confidence Intervals

Compute approximate CIs using the normal approximation (estimate ± 1.96 × SE).

```{r}
if (exists("herit")) {
  herit <- herit %>%
    mutate(
      lower_ci = h2 - 1.96 * SE,
      upper_ci = h2 + 1.96 * SE
    )
  print(herit)
}
```

**Expectation:** The true h² = 0.4 should fall within the 95% CI for most ancestry-method combinations.

## Forest Plot

Visualize h² estimates and CIs across ancestries and methods. The horizontal dashed line marks the ground-truth value of 0.4.

```{r}
if (exists("herit")) {
  ggplot(herit, aes(x = reorder(Ancestry, h2), y = h2, color = method)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
    geom_hline(yintercept = 0.4, linetype = "dashed", color = "gray50", linewidth = 1) +
    labs(
      title = "SNP Heritability Estimates per Ancestry",
      subtitle = "Ground truth h² = 0.4 (dashed line)",
      x = "Ancestry",
      y = expression("Heritability (h"^2*")"),
      color = "Method"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

**Expectation:** Estimates should cluster around 0.4. Confidence intervals should overlap 0.4 for most ancestries. If a method gives consistently biased results, it will be visually apparent here.

## Significant Heritability

Which ancestry-specific h² estimates are statistically significant (p < 0.05)?

```{r}
if (exists("herit")) {
  sig_herit <- herit %>%
    filter(p_value < 0.05) %>%
    arrange(p_value)
  print(sig_herit)
}
```

**Expectation:** Because h² = 0.4 is substantially above zero, most or all ancestry-method combinations should be significant. Non-significant results would indicate insufficient sample size or excessive standard errors for that ancestry group.

## GRM Heatmap

Visualize the genetic relationship matrix to confirm that KING filtering removed related samples. A well-filtered GRM should be near-diagonal.

```{r}
grm_id_path <- file.path(OUT_DIR, "relatedness", "f1.b38.ldpruned.unrelated.grm.id")
grm_bin_path <- file.path(OUT_DIR, "relatedness", "f1.b38.ldpruned.unrelated.grm.bin")
if (file.exists(grm_bin_path) && file.exists(grm_id_path)) {
  grm_ids <- read_tsv(grm_id_path, col_names = c("FID", "IID"))
  n_ids <- nrow(grm_ids)
  grm_bin <- readBin(grm_bin_path, what = "numeric", n = n_ids * (n_ids + 1) / 2)

  grm_mat <- matrix(0, n_ids, n_ids)
  idx <- 1
  for (i in 1:n_ids) {
    for (j in 1:i) {
      grm_mat[i, j] <- grm_bin[idx]
      grm_mat[j, i] <- grm_bin[idx]
      idx <- idx + 1
    }
  }

  n_subsample <- min(200, n_ids)
  set.seed(42)
  subsample_idx <- sample(seq_len(n_ids), n_subsample)

  grm_sub <- grm_mat[subsample_idx, subsample_idx]

  grm_sub %>%
    as_tibble() %>%
    rowid_to_column("row") %>%
    pivot_longer(cols = -row, names_to = "col_str", values_to = "GRM") %>%
    mutate(col = as.integer(str_remove(col_str, "V"))) %>%
    ggplot(aes(x = col, y = row, fill = GRM)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    labs(
      title = "Genetic Relationship Matrix (Subset)",
      subtitle = paste0(n_subsample, " samples — diagonal should dominate after KING filtering"),
      x = "Sample",
      y = "Sample"
    ) +
    theme_minimal() +
    theme(axis.text = element_blank())
}
```

**Expectation:** The diagonal (self-relationship = ~1) should dominate. Off-diagonal values should cluster near 0. If you see substantial off-diagonal structure, KING filtering may not have been aggressive enough, or the GRM was computed before unrelated extraction.

## Per-Ancestry Heritability Comparison

Is h² significantly different between ancestries? Test pairwise.

```{r}
if (exists("herit")) {
  h2_wide <- herit %>%
    filter(method == "mash") %>%
    select(Ancestry, h2, SE)

  if (nrow(h2_wide) >= 2) {
    for (i in 1:(nrow(h2_wide) - 1)) {
      for (j in (i + 1):nrow(h2_wide)) {
        diff_est <- h2_wide$h2[i] - h2_wide$h2[j]
        diff_se <- sqrt(h2_wide$SE[i]^2 + h2_wide$SE[j]^2)
        z <- diff_est / diff_se
        p <- 2 * pnorm(-abs(z))
        cat(sprintf("%s vs %s: Δh² = %.3f ± %.3f, p = %.4f\n",
          h2_wide$Ancestry[i], h2_wide$Ancestry[j], diff_est, diff_se, p))
      }
    }
  }
}
```

**Expectation:** With the same underlying simulated phenotype (h² = 0.4), different ancestries should give similar h² estimates. Significant differences would indicate ancestry-specific genetic architecture or technical artifacts (differences in genotyping quality, sample size, or LD structure).

## Check Your Understanding

1. **The ground-truth h² is 0.4, but the pipeline estimates 0.35.** Is this bias or just sampling variation? How would you decide?
2. **Look at your GRM heatmap. Are there any off-diagonal samples with GRM > 0.05?** What would you do with them if you found them?
3. **Why does LD pruning matter more for heritability estimation than for global ancestry classification?** (Hint: think about the GRM.)

## Troubleshooting

| Problem | Likely Cause | Solution |
|---------|-------------|----------|
| `mash_output.csv` is empty | MASH failed to converge | Check `run_snpHerit` log; try increasing EM iterations or check for collinear covariates |
| GRM heatmap shows strong off-diagonal structure | KING filtering was skipped or pi-hat threshold too lenient | Verify `relatedness: king: true` in config; check the `unrelated_ids.txt` file size |
| h² estimate > 1 (or negative) | REML estimate on boundary; small sample size | With N < 100, REML is unreliable; interpret h² cautiously |
| Different ancestries give very different h² | Possible genuine ancestry-specific architecture, or QC differences | Check if all ancestry groups had the same variant filters applied; check per-ancestry sample sizes |

## Going Further

- **Power analysis**: Given your estimated h² and sample size per ancestry, calculate the minimum detectable effect size for a GWAS at 80% power (using `pwr` package in R).
- **Compare GCTA-REML to MASH**: If the config runs both methods, compute the correlation between their h² estimates. Which has smaller SEs?
- **Simulate a null phenotype**: Generate a phenotype with h² = 0, re-run through heritability estimation, and verify that the pipeline correctly estimates h² ≈ 0 with non-significant p-values.
