---
title: "Lab: Visualizing and Working with Global Ancestry Outputs"
author: "GDCGenomicsQC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab: Visualizing and Working with Global Ancestry Outputs (R + Tidyverse)

This lab covers interacting with output files from the **Global Ancestry Classification Pipeline** using R and tidyverse. You will work with posterior probabilities, ancestry classifications, and confusion matrices.

**Estimated time**: 45 minutes

**Prerequisites**:
- Completed the ancestry classification tutorial (outputs in `OUT_DIR/01-globalAncestry/`)
- R (≥4.0) with tidyverse installed
- Set `OUT_DIR` to your pipeline output directory

---

# Setup: Load Libraries and Set Paths

```{r load-libraries}
library(tidyverse)
OUT_DIR <- "/path/to/your/pipeline/output"
global_anc_dir <- file.path(OUT_DIR, "01-globalAncestry")
```

---

# Section 1: Global Ancestry Output Files Reference

| File Path | Description |
|----------|------------|
| `01-globalAncestry/posterior_probabilities.tsv` | Per-sample posterior probabilities for each ancestry |
| `01-globalAncestry/sample_coords.tsv` | Sample coordinates in PCA/UMAP space |
| `01-globalAncestry/ref_coords.tsv` | Reference coordinates in PCA/UMAP space |
| `01-globalAncestry/ancestry_classifications.tsv` | Predicted ancestry and confidence per sample |
| `01-globalAncestry/keep_{ANC}.txt` | Sample IDs passing ancestry threshold for each ancestry |
| `01-globalAncestry/posterior_probability_stacked_{model}.svg` | Pre-rendered stacked posterior probability plot (model = pca/umap) |

---

# Section 2: Load and Inspect Global Ancestry Data

```{r load-data}
# Load posterior probabilities (PCA model example)
post_prob_path <- file.path(global_anc_dir, "posterior_probabilities.tsv")
if (file.exists(post_prob_path)) {
  post_prob <- read_tsv(post_prob_path)  # Columns: IID, pca_AFR, pca_AMR, pca_EUR, pca_SAS
  glimpse(post_prob)
}

# Load ancestry classifications
class_path <- file.path(global_anc_dir, "ancestry_classifications.tsv")
if (file.exists(class_path)) {
  class <- read_tsv(class_path)  # Columns: IID, pca_predicted, pca_confidence
  glimpse(class)
}

# Join and pivot datasets for visualization
if (exists("post_prob") && exists("class")) {
  anc_data <- post_prob %>%
    left_join(class, by = "IID") %>%
    pivot_longer(
      cols = starts_with("pca_"),
      names_to = "ancestry",
      values_to = "posterior_prob"
    ) %>%
    mutate(ancestry = gsub("pca_", "", ancestry))
  glimpse(anc_data)
}

# Load confusion matrix (if exists)
conf_path <- file.path(global_anc_dir, "ancestry_confusion_matrix.tsv")
if (file.exists(conf_path)) {
  conf_mat <- read_tsv(conf_path)
  print(conf_mat)
}
```

---

# Section 3: Visualize Global Ancestry

```{r visualize-ancestry}
# 1. Stacked posterior probability plot (matches pipeline output)
if (exists("anc_data")) {
  ggplot(anc_data, aes(x = reorder(IID, posterior_prob), y = posterior_prob, fill = ancestry)) +
    geom_col(width = 1) +
    labs(
      title = "Per-Sample Posterior Probabilities for Global Ancestry (PCA)",
      x = "Sample ID",
      y = "Posterior Probability",
      fill = "Ancestry"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_blank(), panel.grid = element_blank())
}

# 2. Posterior probability distribution per ancestry
if (exists("anc_data")) {
  ggplot(anc_data, aes(x = posterior_prob, fill = ancestry)) +
    geom_density(alpha = 0.5) +
    labs(
      title = "Posterior Probability Density per Ancestry",
      x = "Posterior Probability",
      y = "Density",
      fill = "Ancestry"
    ) +
    theme_minimal()
}

# 3. Confusion matrix heatmap (if reported race available)
if (exists("conf_mat")) {
  ggplot(conf_mat, aes(x = predicted, y = reported, fill = n)) +
    geom_tile() +
    geom_text(aes(label = n), color = "white", size = 3) +
    labs(
      title = "Ancestry Classification Confusion Matrix",
      x = "Predicted Ancestry",
      y = "Reported Race",
      fill = "Count"
    ) +
    theme_minimal()
}
```

---

# Section 4: Interactive Analysis

```{r interactive-analysis}
# Count samples per predicted ancestry
if (exists("class")) {
  anc_counts <- class %>%
    count(pca_predicted, sort = TRUE)
  print(anc_counts)
}

# Identify uncertain samples (confidence < 0.8 threshold)
if (exists("class")) {
  uncertain <- class %>%
    filter(pca_confidence < 0.8) %>%
    arrange(pca_confidence)
  print(uncertain)
}

# Calculate average posterior probability per ancestry
if (exists("anc_data")) {
  avg_post <- anc_data %>%
    group_by(ancestry) %>%
    summarise(avg_posterior = mean(posterior_prob), .groups = "drop")
  print(avg_post)
}
```

---

# Next Steps

- Compare PCA vs UMAP posterior probability distributions
- Analyze discordance between predicted ancestry and reported race
- Proceed to the heritability visualization lab for heritability outputs