---
title: "R25 Polygenic Risk Score"
author: "Kody DeGolier"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

# Lab Overview

By the end of this lab you should be able to understand how to use the prs_pipeline to get Polygenic Risk Scores and how to interpret your findings.

# 1. Introduction to Polygenic Risk Scores (PRS)

A Polygenic Risk Score (PRS) is a single value that estimates an individual's genetic liability to a specific trait or disease. It is calculated by summing the risk alleles across the genome, weighted by the effect sizes derived from a Genome-Wide Association Study (GWAS).

The PRS Equation For an individual $j$, the score $PRS_j$ is calculated as:$$PRS_j = \sum_{i=1}^{n} \hat{\beta}_i G_{ij}$$Where:

-   $\hat{\beta}_i$: The effect size (weight) of SNP $i$ from the discovery GWAS.
-   $G_{ij}$: The dosage (0, 1, or 2) of the risk allele for SNP $i$ in individual $j$.

### The Challenge of Diversity

Most discovery GWAS are performed in European-ancestry populations. When these weights are applied to diverse samples, prediction accuracy often drops due to differences in:

1.  **Linkage Disequilibrium (LD):** The correlation structure between SNPs.

2.  **Allele Frequencies:** A common variant in one population may be rare in another.

3.  **Causal Variants:** The actual functional variant might differ or be tagged differently.

# 2. Pipeline Architecture & Inputs

The `prs_pipeline` is a modular SLURM-based tool designed to automate the alignment of GWAS summary statistics with target genotypes and run multiple PRS algorithms in parallel.

### Required Inputs

To run the pipeline, you need:

1.  **Summary Statistics:** GWAS results containing RSIDs, effect alleles, P-values, and Betas.

    1.  **Summary Statistics File Format** The file needs to have a column header and contain the following columns. If there are multiple listed in `{}` that means that any 1 of the options is recognized but the file cannot contain more than 1 of these overlapping columns.

    ```         
    -   {rsid, rs_id, rsids}

    -   {A1, alt, a1}

    -   {p, pval}

    -   {sebeta, beta_se}

    -   beta

    Chromosome and position are not required — they are pulled from the `.bim` file during alignment. `ref`/`A2` is also derived automatically, so it does not need to be in the input.
    ```

2.  **Target Genotype Files:** PLINK format (`.bed`, `.bim`, `.fam`).

3.  **Configuration File:** A `.conf` file defining paths and parameters.

    ```         
    # my_project.conf 
    # --- Paths ---
    summary_stats_file="/scratch.global/GDC/r25outputs/toy/prs_inputs/AFR_EUR/gwas/training_sumstats_singlePRS.txt"
    bim_file_path="/scratch.global/baron063/R25_testing/EUR_simulation_single_pheno_train.bim"
    study_sample="/scratch.global/baron063/R25_testing/EUR_simulation_single_pheno_train"
    output_path="/scratch.global/baron063/r25outputs/toy"
    path_repo="/scratch.global/baron063/prs_pipeline"
    skip_ss_generation=0
    n_total_gwas=300
    ld_matrix_dir="${output_path}/prs_pipeline/ld_matrix"
    gwas_pca_eigenvec_file=""
    phenotype_info_file="/scratch.global/baron063/R25_testing/sample_pheno.txt"
    test_sample="/scratch.global/baron063/R25_testing/EUR_simulation_single_pheno_val"

    # --- Parameters ---

    RUN_CT=true
    RUN_LDPRED2=true
    RUN_LASSOSUM2=true
    RUN_PRSice2=true
    ```

The configuration file above shows the key parameters you will work with. Note the `n_total_gwas` parameter.

**Question:** Why do Bayesian methods like LDpred2 need the total sample size of the discovery GWAS? *Answer: To distinguish between a small effect size and a lack of statistical power in the discovery set.*

### Running the Pipeline

Once your config file is ready, use `sandbox_singularity_runner.sh` to execute the pipeline inside a pre-built Singularity container. Full documentation is at `docs/sandbox_singularity_runner.md` in the `prs_pipeline` repo.

**1. Pull the container image (already done for you — shown for reference):**

``` bash
apptainer pull oras://ghcr.io/mainsqu33ze/gdcgenomicsqc/prsv2:latest
```

**2. Run locally (sandbox mode — for testing):**

``` bash
bash /path/to/prs_pipeline/sandbox_singularity_runner.sh --C my_project.conf -c -l -s
```

**3. Submit to SLURM (production):**

``` bash
sbatch /path/to/prs_pipeline/sandbox_singularity_runner.sh --C my_project.conf -c -l -s -P
```

| Flag | Method                  |
|------|-------------------------|
| `-c` | Clumping + Thresholding |
| `-l` | LDpred2                 |
| `-s` | lassosum2               |
| `-P` | PRSice-2                |
| `-B` | Binary phenotype flag   |

Expected runtime for toy data (3 chromosomes, \~500 samples): \~10–20 min locally, \~5–10 min on SLURM. Full-scale runs with all chromosomes and 10K+ samples can take several hours.

# 3. Comparing PRS Methodologies

The pipeline implements four major single ancestry methods. Each handles **LD (Linkage Disequilibrium)** differently:

|  |  |  |
|------------------------|------------------------|------------------------|
| **Method** | **Approach** | **LD Handling** |
| **C+T** | Heuristic | Physical/Correlation Clumping (removes correlated SNPs) |
| **PRSice-2** | Optimized C+T | Automated High-Resolution Clumping across P-value thresholds |
| **LDpred2** | Bayesian | Models LD using a reference panel and applies a Gibbs sampler |
| **lassosum2** | Penalized Regression | Uses Elastic Net / Coordinate Descent to account for LD |

# 4. Hands-on: Analyzing Pipeline Outputs

Once the pipeline has completed on the HPC, it generates a structured directory. In this section, we will load the scores generated by different methods and compare them.

## Output Structure

The pipeline automatically organizes results into a clean directory tree:

Each method outputs results to `{output_path}/prs_pipeline/`:

| **Method** | **Key Output File**                        |
|------------|--------------------------------------------|
| C+T        | `CT/CT_prs_results.txt`                    |
| LDpred2    | `LDpred2/prs_method_individual_scores.txt` |
| lassosum2  | `lassosum2/prs_method_final_best_prs.csv`  |
| PRSice-2   | `PRSice2/prs_method/PRSice2_outputs.best`  |

## **PRS Score Columns**

All methods output scores into `test_evaluation/` with a unified `SCORE` column, making comparison straightforward:

| **Method** | **Test Evaluation File**                  | **Score Column** |
|------------|-------------------------------------------|------------------|
| C+T        | `test_evaluation/CT_scores.txt`           | `SCORE`          |
| LDpred2    | `test_evaluation/LDpred2_grid_scores.txt` | `SCORE`          |
| lassosum2  | `test_evaluation/lassosum2_scores.txt`    | `SCORE`          |
| PRSice-2   | `test_evaluation/PRSice2_scores.txt`      | `SCORE`          |

### Step 4.1: Load the Data

Replace `output_path` with the directory where your pipeline results are stored.

Code snippet

```{r}
# Load necessary packages
library(data.table)
library(tidyverse)

# Define your output directory
output_path <- "/scratch.global/baron063/r25outputs/toy/prs_pipeline"
test_data_outputs <- file.path(output_path, "test_evaluation") # Important to use the testing data outputs # all others are using the training data so cannot be used for evalution of prs models

# 1. Load Clumping + Thresholding (C+T) Scores
ct_path <- file.path(test_data_outputs, "CT_scores.txt")
ct_data <- fread(ct_path)

# 2. Load LDpred2 Scores
ldp_path <- file.path(test_data_outputs, "LDpred2_grid_scores.txt")
ldp_data <- fread(ldp_path)

# 3. Load lassosum2 Scores
lasso_path <- file.path(test_data_outputs, "lassosum2_scores.txt")
lasso_data <- fread(lasso_path)

# 4. Load PRSice2 Scores (you will add this to the comparison later)
prsice_path <- file.path(test_data_outputs, "PRSice2_scores.txt")
prsice_data <- fread(prsice_path)
```

### Step 4.2: Merge and Correlate

We want to see how much the methods agree on individual risk rankings.

Code snippet

```{r}
# Merge scores by IID so alignment is robust
combined_scores <- merge(ct_data[, c("IID", "SCORE")], ldp_data[, c("IID", "SCORE")],
                         by = "IID", suffixes = c(".ct", ".ld"))
combined_scores <- merge(combined_scores, lasso_data[, c("IID", "SCORE")],
                         by = "IID")
colnames(combined_scores) <- c("IID", "CT", "LDpred2", "lassosum2")
# Hint: merge in PRSice2 the same way to include it in the comparison

# Calculate Pearson Correlation
cor_matrix <- cor(combined_scores[, -1], use = "complete.obs")
print(cor_matrix)

# Visualize the correlation
library(corrplot)
corrplot(cor_matrix, method = "color", addCoef.col = "black")
```

### 4.3: Incorporating Phenotypes & Covariates

To calculate the $R^2$, we need a phenotype file. Usually, this file follows the PLINK format: `FID`, `IID`, `Phenotype`, and several `PC` columns.

R

```{r}
# Load Phenotype and Covariate data
# Assuming a file with columns: FID, IID, Pheno, Age, Sex, PC1, PC2...
pheno_path <- file.path(test_data_outputs, "pheno.txt")
#pheno_path <- "/scratch.global/baron063/R25_testing/sample_pheno.txt"

pheno_data <- fread(pheno_path)
colnames(pheno_data) = c("FID", "IID", "Pheno")

# Merge with our previously created 'combined_scores' data frame
final_data <- merge(combined_scores, pheno_data, by = "IID")

# Ensure the phenotype is numeric
final_data$Pheno <- as.numeric(final_data$Pheno)
```

### 4.4: Calculating Incremental $R^2$

We calculate the $R^2$ by comparing two models:

1.  **Null Model:** Phenotype \~ Age + Sex + PCs

2.  **Full Model:** Phenotype \~ Age + Sex + PCs + **PRS**

The Incremental $R^2$ is $R^2_{Full} - R^2_{Null}$.

```{r}
# Define the base covariates
covariates <- "Age + Sex + PC1 + PC2 + PC3 + PC4 + PC5" # Ideally something like this
covariates <- "Sex"

# Simulate Sex
simulated_sex=rbinom(dim(final_data)[1], 1, 0.5)

# add simulated sex to pheno file
final_data <- cbind(final_data, simulated_sex)
final_data <- final_data %>%
  rename(Sex = simulated_sex)

calculate_incremental_r2 <- function(prs_col, data) {
  # Null Model
  null_formula <- as.formula(paste("Pheno ~", covariates))
  null_model <- lm(null_formula, data = data)
  r2_null <- summary(null_model)$r.squared
  
  # Full Model
  full_formula <- as.formula(paste("Pheno ~", covariates, "+", prs_col))
  full_model <- lm(full_formula, data = data)
  r2_full <- summary(full_model)$r.squared
  
  return(r2_full - r2_null)
}

# Apply to the three methods (add PRSice2 as an exercise)
r2_results <- data.frame(
  Method = c("CT", "LDpred2", "lassosum2"),
  Incremental_R2 = c(
    calculate_incremental_r2("CT", final_data),
    calculate_incremental_r2("LDpred2", final_data),
    calculate_incremental_r2("lassosum2", final_data)
  )
)

print(r2_results)
```

### 4.5: Visualizing Performance Comparison

A simple bar chart is the standard way to show which method "won" for this specific sample.

```{r}
# Visualization using ggplot2
ggplot(r2_results, aes(x = reorder(Method, -Incremental_R2), y = Incremental_R2, fill = Method)) +
  geom_bar(stat = "identity", color = "black", show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PRS Performance Comparison",
    subtitle = "Measured by Incremental R-squared",
    x = "Statistical Method",
    y = "Incremental R2"
  ) +
  scale_fill_brewer(palette = "Set2")
```

### 4.6: Visualizing Predicted vs. Observed (Scatter Plot)

It is also helpful to see the raw correlation for the top-performing method (e.g., CT).

```{r}
# Visualizing the fit for CT
ggplot(final_data, aes(x = CT, y = Pheno)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "darkred") +
  theme_light() +
  labs(
    title = "Observed Phenotype vs. CT Score",
    x = "Polygenic Risk Score (CT)",
    y = "Observed Phenotype Value"
  )
```

**Why is my** $R^2$ so low? In diverse samples, $R^2$ values are often significantly lower than those reported in the original discovery GWAS papers (which are usually 100% European). Ask students to consider if the drop is due to **LD decay** (different correlation patterns) or if the **causal variants** themselves have different effect sizes in this specific population.

# 5. Troubleshooting & Best Practices

### Alignment Issues

If your output `gwas/CT_PRSice2_summary_stat_file.txt` is empty:

-   **Check ID Formats:** Do your Summary Stats use `rsIDs` while your `.bim` file uses `chr:pos`?

-   **Check Strand:** Are the alleles aligned (A/T vs C/G)? The pipeline attempts to flip them, but unusual custom chips may require manual intervention.

### Resource Management (SLURM)

-   **Memory:** LDpred2 is memory-intensive. If the job fails, increase `#SBATCH --mem=64g` to `128g`.

-   **Parallelization:** The pipeline runs methods in parallel. Ensure your `-c` (CPUs) flag matches the requirements of your LD reference panel size.

# 6. Conclusion

By the end of this lab, you should understand that a PRS is not a single "truth" but an estimate dependent on the statistical model used. In diverse populations, comparing simple heuristic methods (C+T) with complex modeling (LDpred2/lassosum2) is essential for validating the portability of genetic risk.

**Assignment:**

1.  Plot the distribution (histogram) of the LDpred2 scores.

2.  Add PRSice2 to the `combined_scores` merge and the `r2_results` table. Did it outperform the other methods on your test data?

3.  If phenotypes are available, calculate the $R^2$ for each method. Which one performs best for your specific sample?

    """
