---
title: "R25 Polygenic Risk Score"
author: "Kody DeGolier"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

# Single Ancestry PRS Lab Overview

By the end of this lab you should be able to understand how to use the prs_pipeline to get Polygenic Risk Scores and how to interpret your findings for single ancestry PRS methods.

# 1. Introduction to Polygenic Risk Scores (PRS)

A Polygenic Risk Score (PRS) is a single value that estimates an individual's genetic liability to a specific trait or disease. It is calculated by summing the risk alleles across the genome, weighted by the effect sizes derived from a Genome-Wide Association Study (GWAS).

The PRS Equation For an individual $j$, the score $PRS_j$ is calculated as:$$PRS_j = \sum_{i=1}^{n} \hat{\beta}_i G_{ij}$$Where:

-   $\hat{\beta}_i$: The effect size (weight) of SNP $i$ from the discovery GWAS.
-   $G_{ij}$: The dosage (0, 1, or 2) of the risk allele for SNP $i$ in individual $j$.

### The Challenge of Diversity

Most discovery GWAS are performed in European-ancestry populations. When these weights are applied to diverse samples, prediction accuracy often drops due to differences in:

1.  **Linkage Disequilibrium (LD):** The correlation structure between SNPs.

2.  **Allele Frequencies:** A common variant in one population may be rare in another.

3.  **Causal Variants:** The actual functional variant might differ or be tagged differently.

# 2. Pipeline Architecture & Inputs

The `prs_pipeline` is a modular SLURM-based tool designed to automate the alignment of GWAS summary statistics with target genotypes and run multiple PRS algorithms in parallel.

### Required Inputs

To run the pipeline, you need:

1.  **Summary Statistics:** GWAS results containing RSIDs, effect alleles, P-values, and Betas.

    1.  **Summary Statistics File Format** The file needs to have a column header and contain the following columns. If there are multiple listed in `{}` that means that any 1 of the options is recognized but the file cannot contain more than 1 of these overlapping columns.

    ```         
    -   {rsid, rs_id, rsids}

    -   {A1, alt, a1}

    -   {p, pval}

    -   {sebeta, beta_se}

    -   beta

    Chromosome and position are not required — they are pulled from the `.bim` file during alignment. `ref`/`A2` is also derived automatically, so it does not need to be in the input.
    ```

    Here is an example summary statistics file with a header:

    ```         
    rsid    A1    beta    beta_se    pval
    rs123   A     0.15    0.02       1.2e-8
    rs456   G     0.08    0.03       5.6e-4
    rs789   C    -0.21    0.04       3.1e-12
    ```

2.  **Target Genotype Files:** PLINK format (`.bed`, `.bim`, `.fam`) split into distinct train and test splits.

    To evaluate a PRS fairly, the model must never see the test data during training. You will split your PLINK data into two independent sets:

    -   **Training set (\~70%):** Used to train the PRS model (tune parameters, fit the model).
    -   **Test set (\~30%):** Held out during training, used only for final evaluation.

    The pipeline includes `run_split_data.sh` which creates random train/test partitions. It can produce sample ID lists only (`-N` flag, fast) or full PLINK file subsets.

3.  **Configuration File:** A `.conf` file defining paths and parameters.

    ```         
    # my_project.conf 
    # --- Paths ---
    summary_stats_file="/scratch.global/GDC/r25outputs/toy/prs_inputs/AFR_EUR/gwas/training_sumstats_singlePRS.txt"
    bim_file_path="/scratch.global/baron063/R25_testing/EUR_simulation_single_pheno_train.bim"
    study_sample="/scratch.global/baron063/R25_testing/EUR_simulation_single_pheno_train"
    output_path="/scratch.global/baron063/r25outputs/toy" # Change the labs config output path
    path_repo="/scratch.global/baron063/prs_pipeline"
    skip_ss_generation=0
    n_total_gwas=300
    ld_matrix_dir="${output_path}/prs_pipeline/ld_matrix" # Output directory | or input if previously ran prs_pipeline
    gwas_pca_eigenvec_file=""
    phenotype_info_file="/scratch.global/baron063/R25_testing/sample_pheno.txt"
    test_sample="/scratch.global/baron063/R25_testing/EUR_simulation_single_pheno_val"

    # --- Parameters ---

    RUN_CT=true
    RUN_LDPRED2=true
    RUN_LASSOSUM2=true
    RUN_PRSice2=true
    ```
    
The config above shows the key parameters. Note how `phenotype_info_file` provides external phenotype values (required for test-set evaluation), and `ld_matrix_dir` avoids recomputing LD if previously generated. Most config values are paths you must provide:

    - `summary_stats_file`, `study_sample`, `bim_file_path`, `output_path`, `path_repo` — always required
    - `afreq_file` — required by LDpred2 and lassosum2 (generated by `plink2 --freq`)
    - `phenotype_info_file` — optional for training, needed for test evaluation
    - `ld_matrix_dir` / `ld_cache_dir` — optional, speeds up re-runs
    - `test_sample` — optional but required for generating test evaluation outputs

    What gets auto-generated: aligned summary statistics, training sample phenotypes, LD matrices (first run), and PRS scores.


The configuration file above shows the key parameters you will work with. Note the `n_total_gwas` parameter.

**Question:** Why do Bayesian methods like LDpred2 need the total sample size of the discovery GWAS? *Answer: To distinguish between a small effect size and a lack of statistical power in the discovery set.*

### Running the Pipeline

Split your genotype data into training and testing sets:

``` bash
bash /path/to/prs_pipeline/run_split_data.sh \
    -1 /path/to/your/study_sample \
    -r /path/to/prs_pipeline \
    -t 70 -T 30
```

After splitting, generate the actual PLINK files and allele frequencies for the training set:

``` bash
plink --bfile /path/to/study_sample \
    --keep /path/to/study_sample/randomization_ids_anc1/train_samples.txt \
    --make-bed --out /path/to/train

plink2 --bfile /path/to/train --freq --out /path/to/train
```

**Assignment:** Generate your own split of the data using `run_split_data.sh` with a 70/30 train/test ratio. Use the sample ID files to extract the corresponding PLINK subsets, then compute allele frequencies for the training set.

Once you have distinct training and testing splits of your genomic data you are ready to build your config file. To build a config file use the template shown above or make a text file and add in these key variables replacing with the appropriate paths. If you are missing any of these please see the Full documentation at `docs/sandbox_singularity_runner.md` in the `prs_pipeline` repo. \* summary_stats_file="/full/path_to_summary_stats_file" \* study_sample="/full_path/genomics_plink/training_split_file_excluding_extension[.bed/.bim/.fam]" \* bim_file_path=same path as above just with the .bim extension \* test_sample="/full_path/genomics_plink/testing_split_file_excluding_extension[.bed/.bim/.fam]" \* path_repo="full_path/prs_pipeline/github_repo" \* output_path="/scratch.global/baron063/r25outputs/toy" \# Change the labs config output path \* n_total_gwas= The number of individuals included in the summary stats generation. \* phenotype_info_file="/scratch.global/baron063/R25_testing/sample_pheno.txt" \# This file contains FID IID pheno. \* ld_matrix_dir="\${output_path}/prs_pipeline/ld_matrix" \# **Optional**; Output directory \| or input if previously ran prs_pipeline \* gwas_pca_eigenvec_file="" \# **Optional**; can be found as an output of the GDCQC pipeline.

Once your config file is ready, use `sandbox_singularity_runner.sh` to execute the pipeline inside a pre-built Singularity container. Full documentation is at `docs/sandbox_singularity_runner.md` in the `prs_pipeline` repo.

**1. Pull the container image (already done for you — shown for reference):**

``` bash
apptainer pull oras://ghcr.io/mainsqu33ze/gdcgenomicsqc/prsv2:latest
```

**2. Run locally (sandbox mode — for testing):**

``` bash
bash /path/to/prs_pipeline/sandbox_singularity_runner.sh --C my_project.conf -c -l -s
```

**3. Submit to SLURM (production):**

``` bash
sbatch /path/to/prs_pipeline/sandbox_singularity_runner.sh --C my_project.conf -c -l -s -P
```

| Flag | Method                  |
|------|-------------------------|
| `-c` | Clumping + Thresholding |
| `-l` | LDpred2                 |
| `-s` | lassosum2               |
| `-P` | PRSice-2                |
| `-B` | Binary phenotype flag   |

Expected runtime for toy data (3 chromosomes, \~500 samples): \~10–20 min locally, \~5–10 min on SLURM. Full-scale runs with all chromosomes and 10K+ samples can take several hours.

**Assignment:** Run the pipeline using only C+T (`-c`). Compare the size of the output and runtime against running all methods. How does the structure of `prs_pipeline/CT/` differ from the multi-method run?

# 3. Comparing PRS Methodologies

The pipeline implements four major single ancestry methods. Each handles **LD (Linkage Disequilibrium)** differently:

|  |  |  |
|------------------------|------------------------|------------------------|
| **Method** | **Approach** | **LD Handling** |
| **C+T** | Heuristic | Physical/Correlation Clumping (removes correlated SNPs) |
| **PRSice-2** | Optimized C+T | Automated High-Resolution Clumping across P-value thresholds |
| **LDpred2** | Bayesian | Models LD using a reference panel and applies a Gibbs sampler |
| **lassosum2** | Penalized Regression | Uses Elastic Net / Coordinate Descent to account for LD |

# 4. Hands-on: Analyzing Pipeline Outputs

Once the pipeline has completed on the HPC, it generates a structured directory. In this section, we will load the scores generated by different methods and compare them.

## Output Structure

The pipeline automatically organizes results into a clean directory tree:

Each method outputs results to `{output_path}/prs_pipeline/`:

| **Method** | **Key Output File**                        |
|------------|--------------------------------------------|
| C+T        | `CT/CT_prs_results.txt`                    |
| LDpred2    | `LDpred2/prs_method_individual_scores.txt` |
| lassosum2  | `lassosum2/prs_method_final_best_prs.csv`  |
| PRSice-2   | `PRSice2/prs_method/PRSice2_outputs.best`  |

## **PRS Score Columns**

All methods output scores into `test_evaluation/` with a unified `SCORE` column, making comparison straightforward:

| **Method** | **Test Evaluation File**                  | **Score Column** |
|------------|-------------------------------------------|------------------|
| C+T        | `test_evaluation/CT_scores.txt`           | `SCORE`          |
| LDpred2    | `test_evaluation/LDpred2_grid_scores.txt` | `SCORE`          |
| lassosum2  | `test_evaluation/lassosum2_scores.txt`    | `SCORE`          |
| PRSice-2   | `test_evaluation/PRSice2_scores.txt`      | `SCORE`          |

### Step 4.1: Load the Data

Replace `output_path` with the directory where your pipeline results are stored.

Code snippet

```{r}
# Load necessary packages
library(data.table)
library(tidyverse)

# Define your output directory
output_path <- "/scratch.global/baron063/r25outputs/toy/prs_pipeline"
test_data_outputs <- file.path(output_path, "test_evaluation") # Important to use the testing data outputs # all others are using the training data so cannot be used for evalution of prs models

# 1. Load Clumping + Thresholding (C+T) Scores
ct_path <- file.path(test_data_outputs, "CT_scores.txt")
ct_data <- fread(ct_path)

# 2. Load LDpred2 Scores
ldp_path <- file.path(test_data_outputs, "LDpred2_grid_scores.txt")
ldp_data <- fread(ldp_path)

# 3. Load lassosum2 Scores
lasso_path <- file.path(test_data_outputs, "lassosum2_scores.txt")
lasso_data <- fread(lasso_path)

# 4. Load PRSice2 Scores (you will add this to the comparison later)
prsice_path <- file.path(test_data_outputs, "PRSice2_scores.txt")
prsice_data <- fread(prsice_path)
```

### Step 4.2: Merge and Correlate

We want to see how much the methods agree on individual risk rankings.

Code snippet

```{r}
# Merge scores by IID so alignment is robust
combined_scores <- merge(ct_data[, c("IID", "SCORE")], ldp_data[, c("IID", "SCORE")],
                         by = "IID", suffixes = c(".ct", ".ld"))
combined_scores <- merge(combined_scores, lasso_data[, c("IID", "SCORE")],
                         by = "IID")
colnames(combined_scores) <- c("IID", "CT", "LDpred2", "lassosum2")
# Hint: merge in PRSice2 the same way to include it in the comparison

# Calculate Pearson Correlation
cor_matrix <- cor(combined_scores[, -1], use = "complete.obs")
print(cor_matrix)

# Visualize the correlation
library(corrplot)
corrplot(cor_matrix, method = "color", addCoef.col = "black")
```

### 4.3: Incorporating Phenotypes & Covariates

To calculate the $R^2$, we need a phenotype file. Usually, this file follows the PLINK format: `FID`, `IID`, `Phenotype`, and several `PC` columns.

R

```{r}
# Load Phenotype and Covariate data
# Assuming a file with columns: FID, IID, Pheno, Age, Sex, PC1, PC2...
pheno_path <- file.path(test_data_outputs, "pheno.txt")
#pheno_path <- "/scratch.global/baron063/R25_testing/sample_pheno.txt"

pheno_data <- fread(pheno_path)
colnames(pheno_data) = c("FID", "IID", "Pheno")

# Merge with our previously created 'combined_scores' data frame
final_data <- merge(combined_scores, pheno_data, by = "IID")

# Ensure the phenotype is numeric
final_data$Pheno <- as.numeric(final_data$Pheno)
```

### 4.4: Calculating Incremental $R^2$

We calculate the $R^2$ by comparing two models:

1.  **Null Model:** Phenotype \~ Age + Sex + PCs

2.  **Full Model:** Phenotype \~ Age + Sex + PCs + **PRS**

The Incremental $R^2$ is $R^2_{Full} - R^2_{Null}$.

```{r}
# Define the base covariates
covariates <- "Age + Sex + PC1 + PC2 + PC3 + PC4 + PC5" # Ideally something like this
covariates <- "Sex"

# Simulate Sex
simulated_sex=rbinom(dim(final_data)[1], 1, 0.5)

# add simulated sex to pheno file
final_data <- cbind(final_data, simulated_sex)
final_data <- final_data %>%
  rename(Sex = simulated_sex)

calculate_incremental_r2 <- function(prs_col, data) {
  # Null Model
  null_formula <- as.formula(paste("Pheno ~", covariates))
  null_model <- lm(null_formula, data = data)
  r2_null <- summary(null_model)$r.squared
  
  # Full Model
  full_formula <- as.formula(paste("Pheno ~", covariates, "+", prs_col))
  full_model <- lm(full_formula, data = data)
  r2_full <- summary(full_model)$r.squared
  
  return(r2_full - r2_null)
}

# Apply to the three methods (add PRSice2 as an exercise)
r2_results <- data.frame(
  Method = c("CT", "LDpred2", "lassosum2"),
  Incremental_R2 = c(
    calculate_incremental_r2("CT", final_data),
    calculate_incremental_r2("LDpred2", final_data),
    calculate_incremental_r2("lassosum2", final_data)
  )
)

print(r2_results)
```

### 4.5: Visualizing Performance Comparison

A simple bar chart is the standard way to show which method "won" for this specific sample.

```{r}
# Visualization using ggplot2
ggplot(r2_results, aes(x = reorder(Method, -Incremental_R2), y = Incremental_R2, fill = Method)) +
  geom_bar(stat = "identity", color = "black", show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "PRS Performance Comparison",
    subtitle = "Measured by Incremental R-squared",
    x = "Statistical Method",
    y = "Incremental R2"
  ) +
  scale_fill_brewer(palette = "Set2")
```

### 4.6: Visualizing Predicted vs. Observed (Scatter Plot)

It is also helpful to see the raw correlation for the top-performing method (e.g., CT).

```{r}
# Visualizing the fit for CT
ggplot(final_data, aes(x = CT, y = Pheno)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "darkred") +
  theme_light() +
  labs(
    title = "Observed Phenotype vs. CT Score",
    x = "Polygenic Risk Score (CT)",
    y = "Observed Phenotype Value"
  )
```

**Why is my** $R^2$ so low? In diverse samples, $R^2$ values are often significantly lower than those reported in the original discovery GWAS papers (which are usually 100% European). Ask students to consider if the drop is due to **LD decay** (different correlation patterns) or if the **causal variants** themselves have different effect sizes in this specific population.

# 5. Troubleshooting & Best Practices

### Alignment Issues

If your output `gwas/CT_PRSice2_summary_stat_file.txt` is empty:

-   **Check ID Formats:** Do your Summary Stats use `rsIDs` while your `.bim` file uses `chr:pos`?

-   **Check Strand:** Are the alleles aligned (A/T vs C/G)? The pipeline attempts to flip them, but unusual custom chips may require manual intervention.

### Resource Management (SLURM)

-   **Memory:** LDpred2 is memory-intensive. If the job fails, increase `#SBATCH --mem=64g` to `128g`.

-   **Parallelization:** The pipeline runs methods in parallel. Ensure your `-c` (CPUs) flag matches the requirements of your LD reference panel size.

# 6. Conclusion

By the end of this lab, you should understand that a PRS is not a single "truth" but an estimate dependent on the statistical model used. In diverse populations, comparing simple heuristic methods (C+T) with complex modeling (LDpred2/lassosum2) is essential for validating the portability of genetic risk.

**Assignment:**

1.  Plot the distribution (histogram) of the LDpred2 scores.

2.  Add PRSice2 to the `combined_scores` merge and the `r2_results` table. Did it outperform the other methods on your test data?

3.  If phenotypes are available, calculate the $R^2$ for each method. Which one performs best for your specific sample?

    """
